---
title: "Formulas"
author: "David Gerard"
date: "`r Sys.Date()`"
output:  
  html_document:
    toc: true
    toc_depth: 4
    toc_float: false
    highlight: pygments
urlcolor: "blue"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo       = TRUE, 
                      fig.align  = "center",
                      fig.height = 3, fig.width = 4)
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(strip.background = ggplot2::element_rect(fill = "white")))
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy()
```

# Learning Objectives

- Overview of Formulas
- [`{lme4}` Vignette](https://cran.r-project.org/web/packages/lme4/vignettes/lmer.pdf)

# Motivation

- A `formula` is a basic data structure in R.

- Whenever you have used a tilde (`~`), you have used a `formula`.

- `formula`s are used in `t.test()`, `lm()`, `glm()`, `lme4::lmer()`, `lme4::nlmer()`, `facet_grid()`, `facet_wrap()`, `purrr::map()` and relatives, `reshape2::acast()`, `aggregate()`, etc...

- We will learn about 
    1. The structure of a `formula`
    2. The syntax for `formula`s.
    3. Functions to work with `formula`s

# Formula Creation

- Formulas are data structures that consist of 
    1. A left-hand side (LHS) `call`
    2. A right-hand side (RHS) `call`
    3. An environment.
    
- So they are two calls with an environment attached (like a quosure).
  - Formally, they are expressions of length 3, where the first element is the function call `~`, the second element is the LHS, and the third element is the RHS. They have a `.Environment` attribute.

- You commonly create a formula with a tilde `~`.

    ```{r}
    f1 <- y ~ x + z
    class(f1)
    typeof(f1) ## built on calls
    attributes(f1)
    length(f1)
    f1[[1]]
    f1[[2]]
    f1[[3]]
    ```

- Test if an object is a formula with `rlang::is_formula()`.

    ```{r}
    rlang::is_formula(f1)
    ```

- The `~` uses the call environment as the formula environment. You can specify other environments using `rlang::new_formula()`.

    ```{r}
    e <- rlang::env()
    rlang::new_formula(lhs = rlang::expr(y), rhs = rlang::expr(x), env = e)
    ```

- You can have one-sided formulas with only a RHS, in which case they have length 2.

    ```{r}
    f2 <- ~ x + y
    f2
    length(f2)
    ```

# Formula Syntax

- In modeling contexts, the LHS contains the name of the response variable. 

- The RHS contains the predictor variables that are linearly related to the response (up to some link function).

- Terms separated by a `+` indicate main effects.

    ```{r}
    mpg ~ wt + drat + hp
    ```
    
    $$
    \text{mpg} = \beta_0 + \beta_1\text{wt} + \beta_2\text{drat} + \beta_3 \text{hp}
    $$
    
- Terms separated by a `:` indicate interaction effects.

    ```{r}
    mpg ~ wt:drat + hp
    ```
    
    $$
    \text{mpg} = \beta_0 + \beta_{12}\text{wt}\times\text{drat} + \beta_3 \text{hp}
    $$

- Terms separated by a `*` indicate interaction and main effects.

    ```{r}
    mpg ~ wt * drat + hp
    ```

    $$
    \text{mpg} = \beta_0 + \beta_1\text{wt} + \beta_2\text{drat} + \beta_{12}\text{wt}\times\text{drat} + \beta_3 \text{hp}
    $$
    
- Note that `a*a` does **not** give you a quadratic term in `a`.

    ```{r}
    lm(mpg ~ wt * wt, data = mtcars)
    ```


- The `^` indicates crossing to a provided degree.

    ```{r}
    mpg ~ (wt + drat + hp)^2
    ```

    is the same as
    ```{r}
    mpg ~ (wt + drat + hp) * (wt + drat + hp)
    ```
    which is the same as
    ```{r}
    mpg ~ wt + drat + hp + wt:drat + wt:hp + drat:hp
    ```

- The `-` operator removes terms.
    ```{r}
    mpg ~ wt*drat + hp - wt
    ```
    is the same as
    ```{r}
    mpg ~ drat + wt:drat + hp
    ```

- The `1` is an implicit variable for the $y$-intercept. You can remove it via subtraction
    ```{r}
    mpg ~ mpg + drat + hp - 1
    ```

    $$
    \text{mpg} = \beta_1\text{wt} + \beta_2\text{drat} + \beta_3 \text{hp}
    $$
    
## `{lme4}`

- Mixed models allow for model parameters to be treated as random variables (e.g. random slopes or intercepts).

- You do this to control for correlation between samples. E.g.
    - Many observations collected on the same individuals.
    - Observations are a part of a hierarchy. E.g. data collected on students who belong to schools who belong to districts.
    
- Lots of folks prefer treating effects as random (as opposed to fixed) when they are interested as the population as a whole more than specific values of a covariate.

- E.g. in a study on sleep deprivation, Researchers measured reaction times for individuals after so many days of sleep deprivation.
    
    ```{r, fig.width = 6, fig.height = 6, message = FALSE}
    library(ggplot2)
    data("sleepstudy", package = "lme4")
    ggplot(sleepstudy, aes(x = Days, y = Reaction)) +
      facet_wrap(. ~ Subject) +
      geom_point() +
      geom_smooth(method = "lm", se = FALSE)
    ```

    
- Each subject has different slopes and intercepts.

- We don't care about the slopes and intercepts for each subject, but we want to account for correlation within subjects.

- So we might consider the slope and intercept of each subject to be random.

- Folks don't just include fixed effects for subject because it would use a lot more degrees of freedom (this is a cynical view).

- To specify just random intercepts (different for each subject), do

    ```{r}
    lm1 <- lme4::lmer(Reaction ~ Days + (1 | Subject), data = sleepstudy)
    ```

- To specify random slopes (different slope on `Day` for each subject), do
    ```{r}
    lm2 <- lme4::lmer(Reaction ~ Days + (Days | Subject), data = sleepstudy)
    ```

- Let 
    - $y_i$ be the reaction time for observation $i$.
    - $x_i$ be the day for observation $i$.
    - $z_{ij}$ be an indicator for subject $j$.
    
- Then `lm1` fit the model
    $$
    y_i = \beta_0 + \beta_1x_i + \sum_{j=1}^{18}\alpha_{0j}z_{ij} + \epsilon_i\\
    \alpha_{0j} \overset{iid}{\sim} N(0, \sigma^2)\\
    \epsilon_i \overset{iid}{\sim} N(0, \tau^2)
    $$
  
- `lm2` fit the model
    $$
    y_i = \beta_0 + \beta_1x_i + \sum_{j=1}^{18}\alpha_{0j}z_{ij} + \sum_{j=1}^{18}\alpha_{1j}x_iz_{ij} + \epsilon_i\\
    (\alpha_{0j},\alpha_{1j}) \overset{iid}{\sim} N_{2}(0, \Sigma)\\
    \epsilon_i \overset{iid}{\sim} N(0, \tau^2)
    $$
  
- The idea is that we specify the random effects with the parentheses. The left of the `|` is the explanatory variable to model a relationship with the response, to the right of the pipe is the grouping factor that is assumed to be random.

- E.g., in `lm1` the the $y$-intercept was different for each subject.

- E.g., in `lm2` the the slope of Day and $y$-intercept was different for each subject.

- Here are the common formulations (Table 2 from `{lme4}` vignette):

    | Formula | Meaning |
    |---------|---------|
    | `y ~ (1 | g)`                                     | Random intercept with fixed mean. |
    | `y ~ (1 | g1/g2)` or `y ~ (1 | g1) + (1 | g1:g2)` | Intercept varying among `g1` and `g2` within `g1`. |
    | `y ~ (1 | g1) + (1 | g2)`                         | Intercept varying among g1 and g2. |
    | `y ~ x + (x | g)`                                 | Correlated random intercept and slope |
    
    - `y` is the response, `x` is the explanatory, and `g` `g1` `g1` are grouping variables.

# Working with Formulas

- `all.vars()` will return a character vector of variable names used in the formula

    ```{r}
    f <- mpg ~ wt:drat + hp
    all.vars(f)
    ```

- Add variables to a formula with `update.formula()`. You use a `.` to represent everything on LHS or RHS.

    ```{r}
    f <- y ~ x
    f <- update.formula(old = f, new = ~ . + z)
    f
    ```

- `model.frame()` will take as input a formula and a data frame and will create a new data frame containing all of the variables involved in the formula. If the variables are not in the data frame, R will look for them in the formula environment.

    ```{r}
    f <- mpg ~ wt:drat + hp
    mfout <- model.frame(formula = f, data = mtcars)
    head(mfout)
    attributes(mfout)
    ```

- The order of the variables are:
    1. The response variable is the first column
    2. All of the predictor variables are in subsequent columns.

- The resulting data frame has a `terms` attribute, which you can extract via `terms()`.
    ```{r}
    terms(mfout)
    ```

- The `terms` object is the supplied formula with attributes that can be used to make the model matrix.

- The `terms` object can be read about via
    ```{r, eval = FALSE}
    ?terms.object
    ```
    
- You can do manipulations on `terms` objects with `delete.response()`, `drop.terms()`, and `reformulate()`. 

- `model.matrix()` can be used to create the design matrix (the $X$ matrix from regression).

    ```{r}
    X <- model.matrix(object = f, data = mtcars)
    head(X)
    ```

- You either provide it a data frame to look for variables, or it looks for variables in the formula environment.

- `model.matrix()` basically evaluates interactions, and includes the intercept.

- `model.matrix()` can also be used on the output of `model.frame()`. Just make sure to include the `data` argument.
    ```{r}
    X <- model.matrix(object = mfout, data = mtcars)
    head(X)
    ```
    
- Use `model.response()` on the output of `model.frame()` to get the response vector matrix.

    ```{r}
    y <- model.response(mfout)
    head(y)
    ```
    
# Using formulas in a function

1. Include `formula` and `data` arguments. 
2. Use `model.frame()`
3. Extract the response vector (`model.response()`) and design matrix (`model.matrix()`) from the model frame.
4. Do what you want with those entities.

- Example:
    ```{r}
    f <- function(formula, data) {
      stopifnot(rlang::is_formula(formula))
      mf <- model.frame(formula = formula, data = data)
      y <- model.response(mf)
      X <- model.matrix(object = mf, data = data)
      ## Continue with your method ...
    }
    ```


- Sometimes you want to treat the different explanatory variables differently. One way would be to (i) separate them by the pipe `|`, (ii) convert to a string, (iii) do string manipulation, (iv) then use `as.formula()` to coerce back to a formula.

    ```{r}
    f <- mpg ~ wt + drat | am + hp
    chvec <- as.character(f)
    xvec <- strsplit(chvec[[3]], "\\|")[[1]]
    f1 <- as.formula(paste0(chvec[[2]], "~", xvec[[1]], "-1"))
    f2 <- as.formula(paste0(chvec[[2]], "~", xvec[[2]], "-1"))
    X1 <- model.matrix(object = f1, data = mtcars, intercept = FALSE)
    X2 <- model.matrix(object = f2, data = mtcars)
    f1
    f2
    head(X1)
    head(X2)
    ```
    
# `{rlang}` Formula Manipulation

- `{rlang}` has a lot of `formula` manipulation tools that aren't useful for working with datasets, but are useful if you want to use `formula`'s as data structures in their own right.

- Create a new formula with `rlang::new_formula()`.
    ```{r}
    fnew <- rlang::new_formula(lhs = rlang::expr(y + w), rhs = rlang::expr(x + z))
    ```

- Test for formula with `rlang::is_formula()`.
    ```{r}
    rlang::is_formula(fnew)
    ```

- Get or set the formula environment with `rlang::f_env()`
    ```{r}
    rlang::f_env(fnew)
    ```
    
- Get or set the LHS with `rlang::f_lhs()`
    ```{r}
    rlang::f_lhs(fnew)
    ```
    
- Get or set the RHS with `rlang::f_rhs()`
    ```{r}
    rlang::f_rhs(fnew)
    ```
    
- Convert RHS to text with `rlang::f_text()`
    ```{r}
    rlang::f_text(fnew)
    ```

# New Functions

- `~` Create a formula using call environment.
- `rlang::new_formula()`: Create a formula using any other environment.
- `rlang::is_formula()`: Predicate for formula.
- `terms.formula()`: Extract terms from a formula.
- `update.formula()`: Add variables to a formula.
- `all.vars()`: Get variable names used in a formula.
- `model.matrix()`: Get design matrix.
- `model.frame()`: Get matrix of variables used in model.
- `model.response()`: Get the response matrix from the output of `model.frame()`.
- `as.formula()`: Coerce to a formula.
