---
title: "R Futures and Parallel Processing"
author: "David Gerard"
date: "`r Sys.Date()`"
output:  
  html_document:
    toc: true
    toc_depth: 4
    toc_float: false
    highlight: pygments
urlcolor: "blue"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo       = TRUE, 
                      fig.align  = "center",
                      fig.height = 3, fig.width = 4)
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(strip.background = ggplot2::element_rect(fill = "white")))
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy()
```

# Learning Objectives

- Futures, batch jobs, parallel processing.
- [Comprehensive Overview of Futures](https://future.futureverse.org/articles/future-1-overview.html)

# Overview

- **future** is the result of an expression that may or may not have been evaluated.

- When you use a future, you evaluate the expression and get the result.

- Futures can be evaluated sequentially or in parallel, on the same machine or on a distributed cluster of machines.

- The idea of a future is that you can write the **same R code** for sequential single computer jobs, as well as for large parallel jobs on the supercomputer, and the R code will work in both scenarios. This is pretty awesome.

# New Functions

