---
title: "S4 and Bioconductor"
author: "David Gerard"
date: "`r Sys.Date()`"
output:  
  html_document:
    toc: true
    toc_depth: 4
    toc_float: false
    highlight: pygments
urlcolor: "blue"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo       = TRUE, 
                      fig.align  = "center",
                      fig.height = 3, fig.width = 4)
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(strip.background = ggplot2::element_rect(fill = "white")))
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy()
```

# Learning Objectives

- [Common Bioconductor Methods and Classes](https://www.bioconductor.org/developers/how-to/commonMethodsAndClasses/)
- [BiocGenerics](https://bioconductor.org/packages/3.14/bioc/html/BiocGenerics.html)
- [S4Vectors](https://bioconductor.org/packages/3.14/bioc/html/S4Vectors.html)
- [IRanges](https://bioconductor.org/packages/3.14/bioc/html/IRanges.html)
- [VariantAnnotation](https://bioconductor.org/packages/release/bioc/html/VariantAnnotation.html)

# Motivation

- The most common application of S4 objects is in the [Bioconductor](https://bioconductor.org/) project, so we will use some of those base classes as an example.

# Bioconductor File Structure

- All class definitions in "R/AllClasses.R"
- All generic function definitions in "R/AllGenerics.R"
- Methods are defined in a file named by the generic function. For example, all `show` methods would go in "R/show-methods.R".


