---
title: "A Brief Overview of Metaprogramming"
author: "David Gerard"
date: "`r Sys.Date()`"
output:  
  html_document:
    toc: true
    toc_depth: 4
    toc_float: false
    highlight: pygments
urlcolor: "blue"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo       = TRUE, 
                      fig.align  = "center",
                      fig.height = 3, fig.width = 4)
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(strip.background = ggplot2::element_rect(fill = "white")))
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy()
```

# Learning Objectives

- Overview of Metaprogramming.
- Learn a couple tricks.
- Required: Chapter 17 of [Advanced R](https://adv-r.hadley.nz/)
- Optional: Hadley goes into a lot of depth in Chapters 18--21 of [Advanced R](https://adv-r.hadley.nz/). But we'll just learn some common and simple use-cases. 
    - Those chapters only need to be understood in great depth if you actually want to help contribute to the tidyverse.

# Overview

- Metaprogramming refers to treating code as data. You can save code as a variable, modify the code with other code, then evaluate the code.

- In R, we call saved code **expressions**.

- You can use `rlang::expr()` to create an expression.

    ```{r}
    my_exp <- rlang::expr(x + y)
    ```
    
- The result is an expression of type `language`
    ```{r}
    typeof(my_exp)
    ```

- If we print out `my_exp`, we just see our code
    ```{r}
    my_exp
    ```
- We can **evaluate** it (make the code run and provide a result) with `base::eval()`

    ```{r}
    x <- 10
    y <- 11
    base::eval(my_exp)
    ```

- The main ideas of metaprogramming:
  1. Treat source code as data
  2. Possibly modify source code programattically,
  3. Evaluate source code in a given environment.
  
- R is one of the few popular languages that provides extensive support for metaprogramming, another one being [Lisp](https://en.wikipedia.org/wiki/Lisp_(programming_language)) (and its children).

- We will use `{rlang}` for manipulating expressions.
    
    ```{r}
    library(rlang)
    ```
  
# Types of Expressions

- When **constants** are captured, they are just returned as the constant that they are.

    ```{r}
    x <- rlang::expr(2)
    typeof(x)
    y <- rlang::expr("two")
    typeof(y)
    ```

- **Symbols** are names of objects.

    ```{r}
    y <- rlang::expr(x)
    x <- c(22, 11, 13)
    eval(y)
    typeof(y)
    ```

- **Calls** are function calls. Base R calls these **languages**.
    ```{r}
    y <- rlang::expr(sum(x))
    x <- c(22, 11, 13)
    eval(y)
    typeof(y)
    ```
    - Calls act a lot like lists, where you can access arguments with list-like operators.
        ```{r}
        fun <- rlang::expr(f(x, y))
        fun
        ```
    - The first element of a call is the function argument.
        ```{r}
        fun[[1]]
        ```
    - Each additional element of a call is an argument.
        ```{r}
        fun[[2]]
        fun[[3]]
        ```
    - You can add new arguments by inserting *symbols*.
        ```{r}
        fun[[4]] <- rlang::expr(z)
        fun
        ```
    - You can make arguments have default values.
        ```{r}
        fun$w <- 1
        fun
        ```
  
# Capturing Code

- As we've seen, to capture code, use `rlang::expr()`.

    ```{r}
    x <- 10
    y <- 11
    myexp <- rlang::expr(x + y)
    myexp
    eval(expr = myexp)
    ```
    
- Capturing an expression directly from code is called **quoting** (e.g. from `rlang::expr()`).

- Converting a string to an expression is called **parsing** and can be done with `rlang::parse_expr()`.

    ```{r}
    my_exp2 <- rlang::parse_expr("x + y")
    eval(my_exp2)
    ```
    
- Converting code to a string is called **deparsing**. You do this with `rlang::expr_text()`
    ```{r}
    rlang::expr_text(my_exp2)
    ```
    
- There is **unquoting**, where you selectively evaluate part of the expression. You use `!!` (pronounced "bang-bang") to unquote. So below, `x` is unquoted.

    ```{r}
    x <- 10
    y <- 11
    rlang::expr(y + !!x)
    ```
    
    - Though, `!!` is an invention of `{rlang}`, so if you use base R expression methods it won't work.

- To capture function arguments as an expression, you have to use `rlang::enexpr()` (for "enrich expr()").

    ```{r}
    capture_exp <- function(x) {
      newexp <- rlang::enexpr(arg = x)
      return(newexp)
    }
    abc <- capture_exp(a + b + c)
    abc
    
    a <- 1
    b <- 2
    c <- 3
    eval(abc)
    ```

- The reason is that if you use `rlang::expr(x)` it will just return the expression with `x`, not with the values of the argument that you pass it.

    ```{r}
    capture_bad <- function(x) {
      newexp <- rlang::expr(expr = x)
      return(newexp)
    }
    capture_bad(a + b + c)
    ```
    
- Use `rlang::enxprs()` to capture a list of all arguments in `...`
    ```{r}
    capall <- function(...) {
      all_args <- rlang::enexprs(...)
      return(all_args)
    }
    capall(x + y, y * 2)
    ```

## Evaluated versus Quoted Arguments

- Evaluated arguments are those whose values are used. E.g. in `sum(x)`, the `x` values are being used, but the variable name `x` is not.

- Quoted arguments are those where the code is used. E.g. in `library(rlang)` we use the code `rlang` and not the value that rlang binds to.

- Lots of dplyr functions use quoted arguments (where they use expressions in terms of variable names).

- **Exercise**: Which are evaluated arguments and which are quoted arguments in the below function calls
    ```{r, eval = FALSE}
    dplyr::filter(mtcars, mpg > 20, wt > 3)
    
    ggplot(mtcars, aes(x = wt, y = mpg)) +
      geom_point(col = "red")
    
    mtcars$mpg
    
    mtcars[["mpg"]]
    ```
    
    ```{block, eval = FALSE, echo = FALSE}
    1. `mtcars` is evaluated. `mpg > 20` and `wt > 3` are both quoted.
    
    2. `mtcars` is evaluated. `wt` and `mpg` are quoted. I think `aes(x = wt, y = mpg)` is evaluated since it works outside of the function call.
    
    3. `mtcars` is evaluated, `mpg` is quoted. To see this, consider `` `$`(mtcars, mpg) ``.
    
    4. Both `mtcars` and `"mpg"` are evaluated. To see this, consider `` `[[`(mtcars, "mpg") ``.
    ```
  

## Quosures

- The `{rlang}` team made a new data object called a **quosure** (quoting closure) that bundles an expression with an environment. This makes it so that you do not have ambiguity when you evaluate an expression in a different environment where you have variable bindings. 

- E.g. the following is presents an issue of the user wanting `a` to be 10 but we have `a` being 1000.

    ```{r}
    fun <- function(x) {
      a <- 1000
      x <- rlang::enexpr(x)
      eval(x)
    }
    a <- 10
    fun(a + 1)
    ```

- Using `rlang::enquo()` will create a "quosure" which also includes the function environment. You cannot use `base::eval()` with a quosure, so `{rlang}` proves `rlang::eval_tidy()`.
    ```{r}
    fun <- function(x) {
      a <- 1000
      x <- rlang::enquo(x)
      rlang::eval_tidy(x)
    }
    a <- 10
    fun(a + 1)
    ```
    
- Quosures are mostly used when you use dots `...` in a function argument and want to use that in conjunction with data masking (where you use a data frame like an environment). This is what most `{dplyr}` functions do. E.g. `dplyr::filer()` uses data masking with the `.data` like an environment and different predicates as the expressions. See the following help file for more information

    ```{r}
    ?`topic-quosure`
    ```
  
# Application: Capturing Argument Names

- If we come up with a cool new plotting function, we might want to have the axis names match the variable names. But this does not occur with the naive approach.

    ```{r}
    plot_red <- function(x, y) {
      plot(x, y, col = "red")
    }
    plot_red(x = mtcars$wt, y = mtcars$mpg)
    ```


- We can do better by using quoted arguments, instead of evaluated arguments.

    ```{r}
    plot_red <- function(x, y) {
      xexpr <- rlang::enexpr(x)
      yexpr <- rlang::enexpr(y)
      plot(x, y, col = "red", 
           xlab = rlang::expr_text(xexpr), 
           ylab = rlang::expr_text(yexpr))
    }
    plot_red(x = mtcars$wt, y = mtcars$mpg)
    ```

- But we can do even better than that by also having the user provide the data

    ```{r}
    plot_red <- function(x, y, data) {
      xexpr <- rlang::enexpr(x)
      yexpr <- rlang::enexpr(y)
      plot(x = data[[xexpr]], 
           y = data[[yexpr]], 
           col = "red", 
           xlab = rlang::expr_text(xexpr), 
           ylab = rlang::expr_text(yexpr))
    }
    plot_red(x = wt, y = mpg, data = mtcars)
    ```

- This works because you can subset by an expression as well
    ```{r}
    mtcars[[rlang::expr(mpg)]]
    ```

# Application: Math Notation in Plots

- R has its own weird syntax for plotting mathematical equations. 

- See the following help file for a full description:

    ```{r, eval = FALSE}
    ?plotmath
    ```
    
- R uses `expression`s to implement mathematical notation in R.
    
- E.g. to write the following in R math

    $$
    \sum_{i=1}^\infty x_i^2 + \beta_1
    $$

    we would first create an expression using the R math notation:
    ```{r}
    mathexp <- rlang::expr(sum(x[i]^2, i == 1, infinity) + beta[1])
    ```

- You can then use that expression in `geom_text()` or `xlab()` or whatever.
    ```{r}
    plot.new()
    text(x = 0.5, y = 0.5, labels = mathexp, cex = 2)
    ```
    

- What if you want to place the estimated slope of a regression on a graph?

    ```{r}
    library(ggplot2)
    pl <- ggplot(mtcars, aes(x = wt, y = mpg)) +
      geom_point() +
      geom_smooth(method = "lm", se = FALSE)
    ```

- First, we calculate the OLS estimates:
    ```{r}
    lmout <- lm(mpg ~ wt, data = mtcars)
    coefvec <- coef(lmout)
    coefvec
    ```

- We could add the equation manually
    ```{r, warning=FALSE, message=FALSE}
    mathexp <- rlang::expr(hat(y)[i] == 37.2 - 5.3 * x[i])
    pl + annotate(geom = "text", x = 4.5, y = 30, label = mathexp)
    ```

- However, this is prone to error (e.g., did you notice that I incorrectly rounded down to 37.2?)

- Just including the coefficient vectors won't work.
    ```{r, warning = FALSE, message = FALSE}
    mathexp <- rlang::expr(hat(y)[i] == coefvec[[1]] - coefvec[[2]] * x[i])
    pl + annotate(geom = "text", x = 4.5, y = 30, label = mathexp)
    ```

- Unquoting to the rescue! Just use `!!`.

    ```{r, warning = FALSE, message = FALSE}
    mathexp <- rlang::expr(hat(y)[i] == !!coefvec[[1]] - !!coefvec[[2]] * x[i])
    pl + annotate(geom = "text", x = 4.5, y = 30, label = mathexp)
    ```
    
# Application: Using the tidyverse in a function

- What if you want to create an awesome `{ggplot2}` function that allows the user to insert a data frame and output a simple scatterplot with red points. We will fill in the skeleton below.

    ```{r}
    plot_red <- function(.data, var1, var2) {

    }
    ```

- If we want the user to not have to use strings for the variable names, then we need to capture those arguments

    ```{r}
    plot_red <- function(.data, var1, var2) {
      var1 <- rlang::enexpr(var1)
      var2 <- rlang::enexpr(var2)
    }
    ```

- But if we were to just put `var1` and `var2` into `{ggplot}`, then `{ggplot}` would interpret that as trying to find a variable literally called `var1` and `var2` in `.data`. So we need to then unquote `var1` and `var2` in `ggplot()`.

    ```{r}
    plot_red <- function(.data, var1, var2) {
      var1 <- rlang::enexpr(var1)
      var2 <- rlang::enexpr(var2)
      ggplot(.data, aes(x = !!var1, y = !!var2)) +
        geom_point(col = "red")
    }
    ```

- This works now

    ```{r, erro = TRUE}
    plot_red(.data = mtcars, var1 = wt, var2 = mpg)
    ```

- Remember that the tidyverse mostly uses quoted arguments, not evaluated arguments, when choosing variable names. So this way of quoting and unquoting is common when coding using tidyverse functions.

- This also shows up in Shiny apps when you use `varSelectInput()`.

# Application: Using `dplyr`-Style Code on New Classes.

- In `eval()`, you can use a `data.frame` as the environment instead of an `environment` object. This is called **data masking**

    ```{r}
    eval(expr = rlang::expr(mpg / wt), envir = mtcars)
    ```

- This allows you to make tidyverse style functions without dependencies on the tidyverse.

- Some folks have explored this [to the extreme](https://github.com/yonicd/bplyr).

- E.g. here is a variation of `dplyr::filter()` that only has a dependency on `{rlang}` (though, maybe not as safe as `dplyr::filter()` for some corner cases)

    ```{r}
    #' @param .data A data.frame.
    #' @param e An expression.
    my_filter <- function(.data, e) {
      e <- rlang::enexpr(arg = e)
      .data[eval(expr = e, envir = .data), , drop = FALSE]
    }
    my_filter(mtcars, mpg > 30)
    ```

- Though, the `{rlang}` folks recommend you use quosures, not environments when doing data masking.

    ```{r}
    #' @param .data A data.frame.
    #' @param e An expression.
    my_filter <- function(.data, e) {
      e <- rlang::enquo(arg = e)
      .data[rlang::eval_tidy(expr = e, data = .data), , drop = FALSE]
    }
    my_filter(mtcars, mpg > 30)
    ```

- You aren't going to be using this filter function in real data analysis. But this style of metaprogramming is useful if you have an S3/S4 object and you want to provide `{dplyr}` style functionality.

- E.g., let's create a class of two data frames where they are connected by a common key

    ```{r}
    setClass(
      Class = "cdf", 
      slots = c(
        df1 = "data.frame", 
        df2 = "data.frame", 
        key = "character"
      ),
      prototype = list(
        df1 = data.frame(),
        df2 = data.frame(),
        key = character()
      )
    )
    
    setValidity(
      Class = "cdf", 
      method = function(object) {
        if (length(object@key) == 0 & ncol(object@df1) == 0 & ncol(object@df2) == 0) {
          return(TRUE)
        } else if (length(object@key) > 1) {
          return("key needs to be length 1")
        } else if (!(object@key %in% names(object@df1))) {
          return("key needs to be a column in df1")
        } else if (!(object@key %in% names(object@df1))) {
          return("key needs to be a column in df2")
        } else {
          return(TRUE)
        }
      }
    )
    
    setMethod(
      f = "show", 
      definition = function(object) {
        str(object@key)
        cat("\n")
        str(object@df1)
        cat("\n")
        str(object@df2)
      }
    )
    ```

- We can try this out on the `nycflights13` dataset.
    ```{r}
    library(nycflights13)
    mydf2 <- new(Class = "cdf", df1 = flights, df2 = planes, key = "tailnum")
    ```

- Let's create a `filter_df2()` method that, when applied, will also remove all rows from `df1` that have the keys that were removed from `df2`

    ```{r}
    setGeneric(name = "filter_df2", def = function(obj, expr) standardGeneric("filter_df2"))
    ```
    
    ```{r}
    setMethod(
      f = "filter_df2", 
      signature = "cdf", 
      definition = function(obj, expr) {
        expr <- rlang::enexpr(expr)
        logvec_df2 <- eval(expr = expr, envir = obj@df2)
        which_key <- obj@df2[[obj@key]][logvec_df2]
        logvec_df1 <- obj@df1[[obj@key]] %in% which_key
        
        obj@df1 <- obj@df1[logvec_df1, , drop = FALSE]
        obj@df2 <- obj@df2[logvec_df2, , drop = FALSE]
        return(obj)
      }
    )
    ```


- Let's try it out

    ```{r}
    filter_df2(obj = mydf2, tailnum == "N105UW")
    ```

# New Functions

- `rlang::expr()`: Create an expression (quote).
- `rlang::enexpr()`: Create an expression of an argument within a function (quote).
- `base::eval()`: Evaluate an expression.
- `rlang::expr_text()`: Convert an expression to text (deparse).
- `!!`: Unquote by evaluating single symbol.
- `rlang::quo()`: Create a quosure.
- `rlang::enquo()`: Create a quosure of an argument in within a function.
- `rlang::eval_tidy()`: Evaluate a quosure.
