---
title: "Rcpp Setup"
author: "David Gerard"
date: "`r Sys.Date()`"
output:  
  html_document:
    toc: true
    toc_depth: 4
    toc_float: false
    highlight: pygments
urlcolor: "blue"
bibliography: "08_bib.bib"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo       = TRUE, 
                      fig.align  = "center",
                      fig.height = 3, fig.width = 4)
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(strip.background = ggplot2::element_rect(fill = "white")))
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy()
```

# Learning Objectives

- Setting up Rcpp
- [Rcpp for Everyone](https://teuder.github.io/rcpp4everyone_en/)
- Chapter 25 of [Advanced R](https://adv-r.hadley.nz/rcpp.html)
- [R's C internals](https://github.com/hadley/r-internals)

# When to Go to C++

- Before you get started, the **two rules of software optimization** are [@jackson1975principles]:
    1. Don't do it.
    2. Don’t do it yet (for experts only).

- You are all expert R programmers. So you should always begin in R.

- When you move to C++, only do so in cases where it will help.

- It helps when *all* three of the following occur:
    1. You use a for-loop over a long vector or a large matrix/array
    2. This for-loop cannot be vectorized (because each iteration depends on previous iterations), and 
    3. Each iteration of the for-loop is relatively fast.

- The above is typically the only reason why you should switch. But as you get more advanced computational skills, you might switch to use data structures implemented in C++ like [heaps](https://en.wikipedia.org/wiki/Heap_(data_structure)) or [binary trees](https://en.wikipedia.org/wiki/Binary_tree) etc...
    - A lot of nice ones are in the [STL](https://en.wikipedia.org/wiki/Standard_Template_Library) or [Boost](https://en.wikipedia.org/wiki/Boost_(C%2B%2B_libraries)) libraries.

# Rcpp

- `{Rcpp}` is a fantastic R package that makes integrating R and C++ almost painless.

- In interactive mode, you can write a C++ function by using `Rcpp::cppFunction()`. You put the source code in quotes.

    ```{r}
    Rcpp::cppFunction(
      '
      void hello_world() {
        Rcpp::Rcout << "Hello World";
      }
      '
    )
    ```

- You can then use that funciton like you would any usual R function.

    ```{r}
    hello_world()
    ```

- More frequently, you have a C++ file (ending in ".cpp") that contains the following:

    ```{Rcpp, eval = FALSE}
    #include <Rcpp.h>
    using namespace Rcpp;
    
    // [[Rcpp::export]]
    void hello_world() {
      Rcpp::Rcout << "Hello World";
    }
    ```
    
    Suppose this file is called "hello.cpp", then you can load it into R using
    
    ```{r, eval = FALSE}
    Rcpp::sourceCpp(file = "hello.cpp")
    ```
    
    and then use `hello_world()` as you would any R function.

- People almost never do either of these in practice. If you are using `{Rcpp}`, it is almost always in an R package.

- `usethis::use_rcpp()` will automatically format an R package to have the appropriate structure for interfacing with `{Rcpp}`.

- You will have to include these lines in your documentation somewhere. I typically include them in the documentation for the package. Make sure you replace `package_name` with the name of the package. Follow instructions from `{usethis}`
    ```{r, eval = FALSE}
    #' @useDynLib package_name, .registration = TRUE
    #' @importFrom Rcpp sourceCpp
    NULL
    ```

- After running `usethis::use_rcpp()`, it will
    1. Add the following lines to your DESCRIPTION file:
    
        ```
        LinkingTo: 
            Rcpp
        Imports: 
            Rcpp
        ```
        
    2. Create a `src` directory.
    
- You place all C++ files in the `src` directory.

- All C++ files end in ".cpp".

- Every file should generally begin with
    ```{Rcpp, eval = FALSE}
    #include <Rcpp.h>
    using namespace Rcpp;
    ```

- The `#include <Rcpp.h>` line basically makes Rcpp available to us in C++. 
    - An R analogy would be including the `{Rcpp}` package in the Imports field of the DESCRIPTION folder. We now have access to the functions in from Rcpp.
    
- The `using namespace Rcpp;` line brings the Rcpp namespace into scope so that we can use functions without declaring that they come from Rcpp. E.g., we can use `Rcout` instead of `Rcpp::Rcout` (just like importing a function in R).

- C++ files should be entirely composed of functions (just like R scripts in packages).

- A C++ function is of the form

    ```
    ftype fname(arg1Type arg1, arg2Type arg2 = default2) {
      // code goes here
      return retval;
    }
    ```

- `ftype` is the type that is returned, such as `int` or `double` or `Rcpp::NumericVector`, etc...

- `fname` is the function name that you choose.

- `arg1Type` and `arg2Type` are the types of the arguments for the function.

- One of these arguments, `arg2` has a default value, `default2`.

- All C++ functions end with `return` followed by the object that is being returned.

- Notice that you don't use assignment for creating C++ functions.

- For example, here I return the sum of two integers:

    ```{Rcpp}
    // [[Rcpp::export]]
    int add2(int x, int y) {
      int sum = x + y;
      return sum;
    }
    ```
    
    ```{r}
    add2(1, 2)
    ```

- Comments in C++ are `//`, not `#`.

- If you want R to have access to an C++ function, you must include the exact line right above the function:

    ```
    // [[Rcpp::export]]
    ```

- If you don't include this, it will still be available to your other C++ functions, but not to R.

- You document C++ functions exactly like R code, except you begin `{roxygen2}` lines with `//'`.

    ```{Rcpp, eval = FALSE}
    //' @title Add two integers
    //' 
    //' @param x Integer 1
    //' @param y Integer 2
    //' 
    //' @return Their sum
    //' 
    //' @author David Gerard
    // [[Rcpp::export]]
    int add2(int x, int y) {
      int sum = x + y;
      return sum;
    }
    ```

- The workflow for working with Rcpp is:
    1. Work on a C++ function.
    2. Run `devtools::load_all()` to compile the function and load it into memory.
        - Or run `devtools::document()` if you changed the documentation.
    3. Test the function.
    4. Rinse and repeat.
    
# References
