---
title: "C++ Basics"
author: "David Gerard"
date: "`r Sys.Date()`"
output:  
  html_document:
    toc: true
    toc_depth: 4
    toc_float: false
    highlight: pygments
urlcolor: "blue"
bibliography: "08_bib.bib"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo       = TRUE, 
                      fig.align  = "center",
                      fig.height = 3, fig.width = 4)
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(strip.background = ggplot2::element_rect(fill = "white")))
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy()
```

```{Rcpp, ref.label=knitr::all_rcpp_labels(), include=FALSE}
```

# Learning Objectives

- [Rcpp for Everyone](https://teuder.github.io/rcpp4everyone_en/)
- Chapter 25 of [Advanced R](https://adv-r.hadley.nz/rcpp.html)
- [R's C internals](https://github.com/hadley/r-internals)


# Basic Differences with R

- Every line of code needs to end with a semicolon `;`.

- Assignment is with `=`, not `<-`.

- Every variable needs to be **declared** ahead of time.

- Comments start with `//`.

- Counting begins at 0 (like in python), not 1. This is the most common source of headaches for R programmers migrating to C++.

- Order of function definitions matter.
    - A function is only available if you have defined it earlier in the file.
    - You can declare functions at the top of a file so that you can use them before the function definition.






