---
title: "C++ Basics"
author: "David Gerard"
date: "`r Sys.Date()`"
output:  
  html_document:
    toc: true
    toc_depth: 4
    toc_float: false
    highlight: pygments
urlcolor: "blue"
bibliography: "08_bib.bib"
---

```{r setup, include=FALSE, message = FALSE}
library(ggplot2)
library(ggthemes)
knitr::opts_chunk$set(echo       = TRUE, 
                      fig.align  = "center",
                      fig.height = 3, fig.width = 4)
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(strip.background = ggplot2::element_rect(fill = "white")))
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy()
```

```{Rcpp, ref.label=knitr::all_rcpp_labels(), include=FALSE}
```

```{Rcpp, eval = FALSE, echo = FALSE}
#include <Rcpp.h>
using namespace Rcpp;
```


# Learning Objectives

- Chapters 1--7 and 30 of [Rcpp for Everyone](https://teuder.github.io/rcpp4everyone_en/)
- Chapter 25 of [Advanced R](https://adv-r.hadley.nz/rcpp.html)
- Learning Objectives:
    - Scalars and scalar operations.
    - Printing output.


# Basic Differences with R

- Every line of code needs to end with a semicolon `;`.

- Assignment is with `=`, not `<-`.

- Every variable needs to be **declared** ahead of time.

- Comments start with `//`.

- Scalars are different from vectors.

- Counting begins at 0 (like in python), not 1. This is the most common source of headaches for R programmers migrating to C++.

- Order of function definitions matter.
    - A function is only available if you have defined it earlier in the file.
    - You can declare functions at the top of a file so that you can use them before the function definition.
    
# Hello World

- You can print out to the console using Rcpp with `Rcpp::Rcout` and the `<<` operator.

    ```{Rcpp, eval = FALSE}
    // [[Rcpp::export]]
    void hello_world_1() {
      Rcpp::Rcout << "Hello World";
    }
    ```

    ```{r}
    hello_world_1()
    ```
    
- The `void` means that our function is not returning anything.

- You can chain multiple outputs together.

    ```{Rcpp, eval = FALSE}
    // [[Rcpp::export]]
    void hello_world_2() {
      Rcpp::Rcout << "Hello" << " World";
    }
    ```
    
    ```{r}
    hello_world_2()
    ```
    
- If you want a new line, use `endl` from the `std` namespace.

    ```{Rcpp, eval = FALSE}
    // [[Rcpp::export]]
    void hello_world_3() {
      Rcpp::Rcout << "Hello" 
                  << std::endl
                  << "World";
                  
    }
    ```

    ```{r}
    hello_world_3()
    ```

- Notice that we use `::` to state the library, just like in R.

- Notice how putting `<<` on different lines did not affect things. This is because the end of the command is wherever the semicolon `;` is.

- In non Rcpp C++, you would use `std::cout` instead of `Rcpp::Rcout`.

- Since C++ is a compiled language, and you cannot use it interactively, you often debug by printing lots of output to the console and seeing if the output is consistent with what you expect. You will use `Rcpp::Rcout` a lot.

# Declaration and Scalar Types

- Let's go through a basic example of the differences between R and C++.

    ```{Rcpp, eval = FALSE}
    // [[Rcpp::export]]
    double add2(double x, double y) {
      double z; // declare z to be an double
      z = x + y;
      return z;
    }
    ```

- In the above function, we
  1. Declare that `add2()` will return a double.
  2. Declare that the arguments `x` and `y` must be doubles.
  3. Declare that `z` is a double.
  4. Assign `z` as `x` plus `y`. This uses `=`.
  5. Return `z`.

- Notice how there is a semicolon after each line (except where curly braces are concerned).

- C++ will error if you forget declarations:

    ``` cpp
    // This will not compile
    // [[Rcpp::export]]
    double add2(double x, double y) {
      z = x + y;
      return z;
    }
    ```

- In C++, **scalars** (elements of length 1) are different from vectors (which contain scalars). 

- The different basic scalars are
    - `bool`: This is either `true` or `false` (all lower case). Same as a length 1 R `logical`.
    - `int`: Same as a length 1 R `integer`.
    - `double`: Same as a length 1 R `double`.
    - `string`: Same as a length 1 R `character`.
    
- Note that in C++, `1` is and `int` and `1.0` is a double.

- Arithmetic operations are similar
    - `+`, `-`, `*`, `/`
    
- In C++, you do powers with `pow()` from the `std` namespace.

    ```{Rcpp, eval = FALSE}
    // [[Rcpp::export]]
    double square(double x) {
      return std::pow(x, 2.0);
    }
    ```

    ```{r}
    square(3)
    ```
    
- Logical operations are the same, but no `&` and `|`
    - `&&` is and
    - `||` is or
    - `!` is not

- The comparison between doubles and ints are the same
    - `==`, `!=`, `>`, `<`, `>=`, `<=`
    
- You can use these in `if-else` statements, which are exactly the same as R:

    ``` cpp
    if (condition) {
      // code
    } else if {
      // code
    } else if {
      // code
    } else {
      // code
    }
    ```

- Let's demonstrate this by creating a C++ version of `sign()`.

    ```{Rcpp, eval = FALSE}
    // [[Rcpp::export]]
    int signC(int x) {
      if (x > 0) {
        return 1;
      } else if (x == 0) {
        return 0;
      } else {
        return -1;
      }
    }
    ```
    
    ```{r}
    signC(-3)
    signC(0)
    signC(11)
    ```
    
- **Exercise**: Write a function called `square2()` that squares a number, but does not use `std::pow()`. Which one is faster?

    ```{block, eval = FALSE, echo = FALSE}
    // [[Rcpp::export]]
    double square2(double x) {
      return x*x;
    }
    ```
    
    ```{r, eval = FALSE, echo = FALSE}
    bench::mark(square(9), square2(9))
    ```
    
    ```{block, eval = FALSE, echo = FALSE}
    `square2()` is just a tiny bit faster.
    ```

# Useful Math Functions for Scalars

- Here are a bunch of the math scalar functions I use pretty regularly.

- Most of these are from the `<cmath>` library from the `std` namespace, or from the `<Rmath>` library from the `R` namespace. 

- If you want to check out those R header files where the functions are declared, type in R:
    ```{r, eval = FALSE}
    R.home("include")
    ```

- You can read about the `<cmath>` library at <https://en.cppreference.com>

- Below, I will write out the return value and arguments like so

    ``` cpp
    type lib::name(type, type)
    ```
## `<cmath>` from `std`

- `double std::min(double a, double b)`: Minimum of two objects (usually scalars) of same type.

- `double std::max(double a, double b)`: Maximum of two objects (usually scalars) of same type.

- `double std::ceil(double x)`: Round a double up. 

- `double double std::floor(double x)`: Round a double down.

- `double std::trunc(double x)`: Round a double closer to zero. So negative numbers up and positive numbers down.

- `double std::round(double x)`: Rounds double to nearest integer. Returns a double.

- `double std::abs(double x)`: Absolute value of a double or int.

- `double std::sqrt(double x)`: Computes $\sqrt{x}$

- `double std::exp(double x)`: Computes $e^x$. 

- `double std::exp2(double x)`: Computes $2^x$

- `double std::pow(double x, double y)`: Computes $x^y$.

- `double std::log(double x)`, `double std::log10(double x)`, `double std::log2(double x)`, logs a double with base $e$, $10$, or $2$ (respectively). 

- `double std::lgamma(double arg)`: Computes natural log of the [gamma function](https://en.wikipedia.org/wiki/Gamma_function). Note that `std::lgamma(x)` is $\log[(x-1)!]$ for any integer $x$, so this can be used to calculate log-factorials.

## `<Rmath>` from `R`

### Math Functions

- `double R::beta(double a, double b)`: Compute the [beta function](https://en.wikipedia.org/wiki/Beta_function)

- `double R::lbeta(double a, double b)`: Compute the log of the [beta function](https://en.wikipedia.org/wiki/Beta_function)

- `double R::choose(double n, double k)`: Compute the [combination](https://en.wikipedia.org/wiki/Combination),  $\binom{n}{k}$

- `double R::lchoose(double n, double k)`: Compute the log of the combination, $\log\left[\binom{n}{k}\right]$.

### Distributions

- There are many distributions available from `<Rmath>`, each is parameterized in standard ways (read about them in [Wikipedia](https://en.wikipedia.org/wiki/Probability_distribution#Common_probability_distributions_and_their_applications)).

- Functions that begin with `d` return the [density](https://en.wikipedia.org/wiki/Probability_density_function) (for continuous distributions) or the [probability mass](https://en.wikipedia.org/wiki/Probability_mass_function) (for discrete distributions).

    ```{r, echo = FALSE, fig.show='hold', out.width="40%"}
    x <- seq(-2, 4, length = 500)
    y <- dnorm(x, mean = 1, sd = 1)
    qplot(x, y, geom = "line", ylab = "f(x)") +
      geom_segment(dat = data.frame(x = 2, xend = 2, y = 0, 
                                    yend = dnorm(x = 2, mean = 1, sd = 1)), 
                   aes(x = x, y = y, xend = xend, yend = yend), lty = 2, col = 2)
    
    xb <- 0:6
    yb <- dbinom(x = xb, size = 6, prob = 0.4)
    color <- rep("A", 7)
    color[[4]] <- "B"
    qplot(x = xb, xend = xb, y = 0, yend = yb, color = color, geom = "segment", lwd = I(2)) +
      scale_color_colorblind() +
      scale_x_continuous(breaks = 0:6) +
      xlab("x") +
      ylab("Pr(x)") +
      theme(legend.position = "none")
    ```
    
- Functions that begin with `p` return the probability for being less than or equal to some value. This is called the [cumulative distribution function](https://en.wikipedia.org/wiki/Cumulative_distribution_function).

    ```{r, echo = FALSE, fig.show='hold', out.width="40%"}
    polydf <- data.frame(x = c(min(x), x[x < 2], 2, 0), 
                         y = c(0, y[x < 2], 0, 0))
    qplot(x, y, geom = "line", ylab = "f(x)") +
      geom_polygon(data = polydf, mapping = aes(x = x, y = y), 
                   fill = "blue", alpha = 1/4)

    color <- rep("A", 7)
    color[1:4] <- "B"
    qplot(x = xb, xend = xb, y = 0, yend = yb, color = color, geom = "segment", lwd = I(2)) +
      scale_color_colorblind() +
      scale_x_continuous(breaks = 0:6) +
      xlab("x") +
      ylab("Pr(x)") +
      theme(legend.position = "none")
    ```

- Functions that begin with `q` return the [quantile](https://en.wikipedia.org/wiki/Quantile). That is, you tell me a probability and I will tell you the value such that the probability of being less than that value is the provided probability.

    ```{r, echo = FALSE, fig.show='hold', out.width="40%"}
    qplot(x, y, geom = "line", ylab = "f(x)") +
      geom_polygon(data = polydf, mapping = aes(x = x, y = y), 
                   fill = "blue", alpha = 1/4) +
      annotate(geom = "text", x = 0.5, y = 0.1, label = "0.8413", color = "black")
    
    qplot(x = xb, xend = xb, y = 0, yend = yb, color = color, geom = "segment", lwd = I(2)) +
      scale_color_colorblind() +
      scale_x_continuous(breaks = 0:6) +
      xlab("x") +
      ylab("Pr(x)") +
      theme(legend.position = "none") +
      annotate(geom = "text", x = 0.75, y = 0.25, label = "0.8208")
    ```

- Functions that begin with `r` provide random draws from a given distribution.

- The common arguments between functions are:
    - `x` an observation of that distribution.
    - `p` a probability.
    - `lt`: `lt = 0` means that `p` is the upper tail probability while `lt = 1` means that `p` is the lower tail probability.
    - `lg`: `lg = 0` means that `p` is the probability while `lg = 1` means that `lg` is the log-probability.

- Below are the most common distributions.

- [Normal Distribution](https://en.wikipedia.org/wiki/Normal_distribution) with mean and standard deviation (*not* variance).
    - In Wikipedia: `mu` = $\mu$, `sigma` = $\sigma$.

    ``` cpp
    double R::dnorm(double x, double mu, double sigma, int lg)              
    double R::pnorm(double x, double mu, double sigma, int lt, int lg)      
    double R::qnorm(double p, double mu, double sigma, int lt, int lg)      
    double R::rnorm(double mu, double sigma)
    ```

- [Gamma Distribution](https://en.wikipedia.org/wiki/Gamma_distribution) with shape and scale parameters. 
    - In Wikipedia: `shp` = $k$, `scl` = $\theta$.

    ``` cpp
    double R::dgamma(double x, double shp, double scl, int lg)     
    double R::pgamma(double x, double alp, double scl, int lt, int lg)
    double R::qgamma(double p, double alp, double scl, int lt, int lg)
    double R::rgamma(double a, double scl)
    ```

- [Beta Distribution](https://en.wikipedia.org/wiki/Beta_distribution) with left and right shape parameters. 
    - In Wikipedia: `a` = $\alpha$, `b` = $\beta$.

    ``` cpp
    double R::dbeta(double x, double a, double b, int lg)         
    double R::pbeta(double x, double p, double q, int lt, int lg)
    double R::qbeta(double a, double p, double q, int lt, int lg)
    double R::rbeta(double a, double b)   
    ```

- [Chi-squared Distribution](https://en.wikipedia.org/wiki/Chi-squared_distribution).
    - In Wikipedia: `df` = $k$.
    
    ``` cpp
    double R::dchisq(double x, double df, int lg)          
    double R::pchisq(double x, double df, int lt, int lg)  
    double R::qchisq(double p, double df, int lt, int lg)  
    double R::rchisq(double df)   
    ```

- [$F$-distribution](https://en.wikipedia.org/wiki/F-distribution) with numerator and denominator degrees of freedom.
    - In Wikipedia: `df1` = $d_1$, `df2` = $d_2$.
    
    ``` cpp
    double R::df(double x, double df1, double df2, int lg)      
    double R::pf(double x, double df1, double df2, int lt, int lg)
    double R::qf(double p, double df1, double df2, int lt, int lg)
    double R::rf(double df1, double df2)                
    ```

- [$t$-distribution](https://en.wikipedia.org/wiki/Student%27s_t-distribution) with degrees of freedom.
    - In Wikipedia: `n` = $\nu$.

    ``` cpp
    double R::dt(double x, double n, int lg)            
    double R::pt(double x, double n, int lt, int lg)        
    double R::qt(double p, double n, int lt, int lg)        
    double R::rt(double n)                      
    ```

- [Binomial Distribution](https://en.wikipedia.org/wiki/Binomial_distribution) with size and success probability parameters.
    - In Wikipedia: `n` = $n$, `p` = $p$

    ``` cpp
    double R::dbinom(double x, double n, double p, int lg)      
    double R::pbinom(double x, double n, double p, int lt, int lg)  
    double R::qbinom(double p, double n, double m, int lt, int lg)  
    double R::rbinom(double n, double p)
    ```


- [Exponential Distribution](https://en.wikipedia.org/wiki/Exponential_distribution) with scale parameter.
    - In Wikipedia: `sl` = $1 / \lambda$ (**not** $\lambda$).
    
    ``` cpp
    double R::dexp(double x, double sl, int lg)     
    double R::pexp(double x, double sl, int lt, int lg)
    double R::qexp(double p, double sl, int lt, int lg)
    double R::rexp(double sl)               
    ```
    
    ```{Rcpp, eval = FALSE}
    // [[Rcpp::export]]
    double exp_demo() {
      return R::dexp(1, 2, 0);
    }
    ```
        
    ```{r}
    exp_demo() ## uses scale
    dexp(x = 1, rate = 1/2, log = FALSE) ## uses rate
    ```

- [Geometric Distribution](https://en.wikipedia.org/wiki/Geometric_distribution) with success probability $p$. This uses the model formulation that counts the number of failures before a success.
    - In Wikipedia: `p` = $p$. 

    ``` cpp
    double R::dgeom(double x, double p, int lg)     
    double R::pgeom(double x, double p, int lt, int lg)
    double R::qgeom(double p, double pb, int lt, int lg)
    double R::rgeom(double p)               
    ```
    
    ```{Rcpp, eval = FALSE}
    // [[Rcpp::export]]
    double g_demo() {
      return R::dgeom(0, 0.7, 0);
    }
    ```
    
    ```{r}
    g_demo() ## 70% prob of a success on the first trial, so 0 failures.
    ```
    
- [Hypergeometric Distribution](https://en.wikipedia.org/wiki/Hypergeometric_distribution)
    - In Wikipedia:
        - `x` = $k$ = Number of white balls drawn.
        - `r` = $K$ = number of white balls in urn.
        - `b` = $N - K$ = number of black balls in urn.
        - `n` = $n$ = number of balls drawn.
    
    ``` cpp
    double R::dhyper(double x, double r, double b, double n, int lg)
    double R::phyper(double x, double r, double b, double n, int lt, int lg)
    double R::qhyper(double p, double r, double b, double n, int lt, int lg)
    double R::rhyper(double r, double b, double n)   
    ```
    
- [Poisson Distribution](https://en.wikipedia.org/wiki/Poisson_distribution) with mean parameter.
    - In Wikipedia: `lb` = $\lambda$.

    ``` cpp
    double R::dpois(double x, double lb, int lg)        
    double R::ppois(double x, double lb, int lt, int lg)
    double R::qpois(double p, double lb, int lt, int lg)
    double R::rpois(double mu)  
    ```
