---
title: "Armadillo"
author: "David Gerard"
date: "`r Sys.Date()`"
output:  
  html_document:
    toc: true
    toc_depth: 4
    toc_float: false
    highlight: pygments
urlcolor: "blue"
bibliography: "08_bib.bib"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo       = TRUE, 
                      fig.align  = "center",
                      fig.height = 3, fig.width = 4)
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(strip.background = ggplot2::element_rect(fill = "white")))
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy()
```

```{Rcpp, ref.label=knitr::all_rcpp_labels(), include=FALSE}
```

```{Rcpp, eval = FALSE, echo = FALSE}
#include <RcppArmadillo.h>
// [[Rcpp::depends(RcppArmadillo)]]
using namespace Rcpp;
using namespace arma;
```

# Learning Objectives

- [Armadillo Docs](http://arma.sourceforge.net/docs.html)
- Learning Objectives:
    - Using Armadillo for linear algebra
    
# Integration with R

- If you need to use matrices or arrays in C++, you should use the Armadillo C++ library.

- It contains fast linear algebra algorithms, as well as a nice interface for working with matrices.

- You can set up an R package to use Armadillo by running the following R:
    ```{r, eval = FALSE}
    usethis::use_rcpp_armadillo()
    ```
    
- This will add `RcppArmadillo` to the `LinkingTo` field in the `DESCRIPTION` file. It will also create `Makevars` files to set compilation settings.

- At the top of each C++ file, you now include the following (instead of just including Rcpp):

    ``` cpp
    #include <RcppArmadillo.h>
    // [[Rcpp::depends(RcppArmadillo)]]
    using namespace Rcpp;
    using namespace arma;
    ```

# Armadillo Vectors

# Armadillo Matrices

# Linear Algebra Operations
