---
title: "C++ Vectors"
author: "David Gerard"
date: "`r Sys.Date()`"
output:  
  html_document:
    toc: true
    toc_depth: 4
    toc_float: false
    highlight: pygments
urlcolor: "blue"
bibliography: "08_bib.bib"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo       = TRUE, 
                      fig.align  = "center",
                      fig.height = 3, fig.width = 4)
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(strip.background = ggplot2::element_rect(fill = "white")))
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy()
```

```{Rcpp, ref.label=knitr::all_rcpp_labels(), include=FALSE}
```

```{Rcpp, eval = FALSE, echo = FALSE}
#include <Rcpp.h>
using namespace Rcpp;
```

# Learning Objectives

- Chapters 8, 10, 19, and 24 of [Rcpp for Everyone](https://teuder.github.io/rcpp4everyone_en/)
- Chapter 25 of [Advanced R](https://adv-r.hadley.nz/rcpp.html)
- [Rcpp Quick Reference Guide](https://dirk.eddelbuettel.com/code/rcpp/Rcpp-quickref.pdf)
- Learning Objectives:
    - Rcpp Vectors
    - For-loops
    - Basic Vector Operations

# Rcpp Vectors

- `{Rcpp}` provides classes of C++ vectors which make programming easier for folks used to R.

- The classes are called `LogicalVector`, `IntegerVector`, `NumericVector`, and `CharacterVector`.

- Note, in regular C++ (without Rcpp), you typically use arrays or `std::vector` objects, like `std::vector<double>`.

- If you have a function that accepts and returns vectors, then you use those types.

    <mark style="background-color: LightCoral">C++</mark>

    ``` cpp
    NumericVector fn(IntegerVector x, bool y) {
      // code to create z
      return z;
    }
    ```

# Creating Vectors

- You create an empty vector of length `n` with

    <mark style="background-color: LightCoral">C++</mark>

    ``` cpp
    NumericVector x(n);
    ```

- You can add default values by including a second argument via

    <mark style="background-color: LightCoral">C++</mark>

    ``` cpp
    NumericVector x(n, 2.0);
    ```
    
- If you want an `IntegerVector` from `0` to `n - 1`, probably the best way is through `Rcpp::seq(0, n-1)` `std::iota()`

    <mark style="background-color: LightCoral">C++</mark>

    ``` cpp
    IntegerVector x = Rcpp::seq(0, n - 1);
    ```

- If you have specific elements you want to create the vector with, use curly braces (requires C++11):

    <mark style="background-color: LightCoral">C++</mark>

    ``` cpp
    NumericVector x = {9.0, 1.2, 4.9};
    ```
    
- For older versions of C++, use `NumericVector::create()`.

    <mark style="background-color: LightCoral">C++</mark>

    ``` cpp
    NumericVector x = NumericVector::create(9.0, 1.2, 4.9);
    ```

- Let's demonstrate these methods

    <mark style="background-color: LightCoral">C++</mark>

    ```{Rcpp, eval = FALSE}
    // [[Rcpp::export]]
    void vec_create() {
      int n = 5;
      
      NumericVector x1(n);
      
      NumericVector x2(n, 2.0);
      
      CharacterVector x3(n, "A");
      
      IntegerVector x4 = Rcpp::seq(0, n - 1);
      
      NumericVector x5 = {9.0, 1.2, 4.9};
      
      NumericVector x6 = NumericVector::create(9.0, 1.2, 4.9);
      
      Rcpp::Rcout << "x1: " << x1 << std::endl
                  << "x2: " << x2 << std::endl
                  << "x3: " << x3 << std::endl
                  << "x4: " << x4 << std::endl
                  << "x5: " << x5 << std::endl
                  << "x6: " << x6 << std::endl;
    }
    ```
    
    <mark style="background-color: lightblue">R</mark>
    
    ```{r}
    vec_create()
    ```

# Accessing Elements

- You get and set individual elements of a vector using brackets `[]`, just like in R.

- Note that in C++, **indexing starts at 0**.

- One more time for emphasis, in **C++ indexing starts at 0**.

    <mark style="background-color: LightCoral">C++</mark>

    ```{Rcpp, eval = FALSE}
    // [[Rcpp::export]]
    void subset_example(NumericVector x) {
      Rcpp::Rcout << "x[0]:" << x[0] << std::endl
                  << "x[1]:" << x[1] << std::endl
                  << "x[2]:" << x[2] << std::endl;
                  
      x[0] = 21;
      x[1] = 22;
      x[2] = 23;
      
      Rcpp::Rcout << "x[0]:" << x[0] << std::endl
                  << "x[1]:" << x[1] << std::endl
                  << "x[2]:" << x[2] << std::endl;
    }
    ```
    
    <mark style="background-color: lightblue">R</mark>
    
    ```{r}
    subset_example(c(14, 38, 29))
    ```

# Rcpp Vector Methods

- A **method** is a function that is attached to an object. In C++, methods are accessed with a `.` (just like python).

- Rcpp vectors have a lot of useful methods.

- `length()` or `size()`: Get total number of elements in the vector. These are equivalent.

    <mark style="background-color: LightCoral">C++</mark>

    ``` cpp
    x.length();
    ```

- `fill()`: Fill all elements of a vector with scalar value.

    <mark style="background-color: LightCoral">C++</mark>

    ``` cpp
    x.fill(2.0);
    ```

- `sort()`: Return a vector that sorts the object.

    <mark style="background-color: LightCoral">C++</mark>

    ``` cpp
    x.sort(false); // ascending order
    x.sort(true); // descending order
    ```

- `begin()` and `end()` returns iterators pointing to the first and last elements of the vector (we'll talk about this later)

- Let's demonstrate these.

    <mark style="background-color: LightCoral">C++</mark>

    ```{Rcpp, eval = FALSE}
    // [[Rcpp::export]]
    void method_example() {
      NumericVector x = {1.0, 3.0, 5.0};
      Rcpp::Rcout << "x            : " << x << std::endl;
      
      Rcpp::Rcout << "x.length()   : " << x.length() << std::endl;
      
      Rcpp::Rcout << "x.sort(false): " << x.sort(false) << std::endl;
      Rcpp::Rcout << "x.sort(true) : " << x.sort(true) << std::endl;
      
      x.fill(2.0);
      Rcpp::Rcout << "x.fill(2.0)  : " << x << std::endl;
    }
    ```
    
    <mark style="background-color: lightblue">R</mark>
    
    ```{r}
    method_example()
    ```

- `NumericVector`s have methods to remove and add values, but [Dirk says](https://stackoverflow.com/questions/19472211/how-to-remove-an-element-in-numericvector-for-a-recursion-using-r-and-rcpp) this is not a good idea. So you should use `std::vector<double>` objects to do this efficiently.

# For-loops

- All for-loops in C++ are of the form

    <mark style="background-color: LightCoral">C++</mark>

    ``` cpp
    for (s1; s2; s3) {
      // code executed each iteration
    }
    ```

    - `s1` is code that is executed once before the for-loop. Usually this declares an integer index, such as `int i = 0`.
    - `s2` is a predicate that, when `true`, allows the for-loop to execute another iteration. Usually, this is `i < n` if you want to go through the for-loop `n` times.
    - `s3` is code that is evaluated at the end of each for-loop. This is usually `i++` to add1 to `i`.
    
- Almost all for-loops you write will look like this

    <mark style="background-color: LightCoral">C++</mark>

    ``` cpp
    for (int i = 0; i < n; i++) {
    
    }
    ```

    - We define `i` as an integer, initializing it to be `0`.
    - We only exit the for-loop if `i >= n`.
    - At the end of each iteration we add `1` to `i`.
    
- **Exercise**: In the above for-loop, if `n = 10`, how many iterations will of the for-loop will run, 9 or 10?
    ```{block, eval = FALSE, echo = FALSE}
    10, because we start at 0.
    ```
    
- Let's recreate `sum()` using Rcpp.

    <mark style="background-color: LightCoral">C++</mark>

    ```{Rcpp, eval = FALSE}
    // [[Rcpp::export]]
    double sum2(NumericVector x) {
      int n = x.length();
      double sval = 0.0;
      for (int i = 0; i < n; i++) {
        sval += x[i];
      }
      return sval;
    }
    ```
    
    <mark style="background-color: lightblue">R</mark>

    ```{r}
    x <- runif(100)
    sum(x)
    sum2(x)
    ```
    
- Here are the microbenchmarks:

    ```{r, echo = FALSE}
    knitr::kable(bench::mark(sum(x), sum2(x))[1:9])
    ```

- In R for-loops, you specify the vector you iterate over. In C++, you specify the exit condition and the code to run at the end of each iteration.

- In effect, this makes `for(i in 1:n)` in R the same as `for(int i = 0; i < n; i++)` in C++.

# Aliasing

- If you assign an Rcpp vector `x` to another object `y` using `=`, then the value of `x` is **not** copied to `y`. Rather, `x` becomes an **alias** for `y`. 

- This means that if you edit `x`, then that will edit `y`. And if you edit `y` then that will edit `x`.

- You can use `Rcpp::clone()` to make a copy.

    <mark style="background-color: LightCoral">C++</mark>

    ```{Rcpp, eval = FALSE}
    // [[Rcpp::export]]
    void copy_example() {
      NumericVector x = {1, 2, 3};
      Rcpp::Rcout << x << std::endl;
      
      // cloning
      NumericVector z = Rcpp::clone(x);
      z[0] = 13;
      Rcpp::Rcout << x << std::endl;
      
      // aliasing
      NumericVector y = x;
      y[1] = 10;
      Rcpp::Rcout << x << std::endl;
    }
    ```
    
    <mark style="background-color: lightblue">R</mark>
    
    ```{r}
    copy_example()
    ```

- This is since `x` is binding to a pointer for the vector, not for the vector itself. So copying `x` to `y` just copies the pointer.

# Missing Values and Infinity

- Infinity is encoded using `R_PosInf` and `R_NegInf`.

- Missing values are encoded using `NA_REAL`, `NA_INTEGER`, `NA_LOGICAL`, and `NA_STRING`.

- You can include these in Rcpp vectors without worrying too much.

    <mark style="background-color: LightCoral">C++</mark>

    ``` cpp
    Rcpp::NumericVector x(10, NA_REAL)
    ```
    
- E.g. if you use Rcpp Sugar then Rcpp will understand how to propagate missing values appropriately.

- But if you try to use these missing values as scalars, you have to be very scared.

- `NA_INTEGER` is  the smallest integer allowed in C++, so `NA_INTEGER + 1` would not longer be considered missing data, e.g.

    <mark style="background-color: lightblue">R</mark>

    ```{r}
    Rcpp::evalCpp("NA_INTEGER")
    Rcpp::evalCpp("NA_INTEGER + 1")
    ```

- Coercing `NA_LOGICAL` to a `bool` will evaluate to `true`, since `bool` does not allow for anything except `true` and `false`.

    <mark style="background-color: lightblue">R</mark>

    ```{r}
    Rcpp::evalCpp("(bool)NA_LOGICAL")
    ```

# Exercises

1. (from Advanced R) For each of the following functions, read the code and figure out what the corresponding base R function is. You might not understand every part of the code yet, but you should be able to figure out the basics of what the function does.

    <mark style="background-color: LightCoral">C++</mark>

    ``` cpp
    double f1(NumericVector x) {
      int n = x.size();
      double y = 0;
    
      for(int i = 0; i < n; ++i) {
        y += x[i] / n;
      }
      return y;
    }
    ```
    
    ```{block, eval = FALSE, echo = FALSE}
    Calculates the mean of `x`. Same as `mean()` in R.
    ```
    
    <mark style="background-color: LightCoral">C++</mark>
    
    ``` cpp
    NumericVector f2(NumericVector x) {
      int n = x.size();
      NumericVector out(n);
    
      out[0] = x[0];
      for(int i = 1; i < n; ++i) {
        out[i] = out[i - 1] + x[i];
      }
      return out;
    }
    ```
    
    ```{block, eval = FALSE, echo = FALSE}
    This calculates the cumulative summation of `x`. Same as `cumsum()` in R.
    ```
    
    <mark style="background-color: LightCoral">C++</mark>
    
    ``` cpp
    bool f3(LogicalVector x) {
      int n = x.size();
    
      for(int i = 0; i < n; ++i) {
        if (x[i]) return true;
      }
      return false;
    }
    ```

    ```{block, eval = FALSE, echo = FALSE}
    This returns `true` if any value in `x` is `true`, and returns `false` otherwise. Same as `any()` in R.
    ```
    
    <mark style="background-color: LightCoral">C++</mark>
    
    ``` cpp
    int f4(Function pred, List x) {
      int n = x.size();
    
      for(int i = 0; i < n; ++i) {
        LogicalVector res = pred(x[i]);
        if (res[0]) return i + 1;
      }
      return 0;
    }
    ```

    ```{block, eval = FALSE, echo = FALSE}
    Takes a predicate function `pred` and a list of elements `x` to apply that function to and returns the first position of `x` that a `true` is returned. Same as `Position()` in R.
    ```
    
    <mark style="background-color: LightCoral">C++</mark>
    
    ``` cpp
    NumericVector f5(NumericVector x, NumericVector y) {
      int n = std::max(x.size(), y.size());
      NumericVector x1 = rep_len(x, n);
      NumericVector y1 = rep_len(y, n);
    
      NumericVector out(n);
    
      for (int i = 0; i < n; ++i) {
        out[i] = std::min(x1[i], y1[i]);
      }
    
      return out;
    }
    ```

    ```{block, eval = FALSE, echo = FALSE}
    Returns element-wise minimum of `x` and `y`. This is the same as `pmin()` in R.
    
    `rep_len()` is an Rcpp sugar function that repeats a vector until it is a certain length.
    ```

2. Create a function in C++ call `fib()` that will take as input `n` and return the first `n` [Fibonacci numbers](https://en.wikipedia.org/wiki/Fibonacci_number), where the sequence begins with 0, 1, 1, 2, 3, 5, 8, 13, ...

    E.g.

    ```{Rcpp, eval = FALSE, echo = FALSE}
    // [[Rcpp::export]]
    IntegerVector fib(int n) {
      IntegerVector out(n);
      
      out[0] = 0;
      if (n == 1) {
        return out;
      }
      
      out[1] = 1;
      if (n == 2) {
        return out;
      }
      
      for (int i = 2; i < n; i++) {
        out[i] = out[i - 1] + out[i - 2];
      }
      
      return out;
    }
    ```
    
    <mark style="background-color: lightblue">R</mark>
    
    ```{r}
    fib(1)
    fib(2)
    fib(10)
    ```


3. (from Advanced R) Convert the following functions into C++. For now, assume the inputs have no missing values. Try not to use Rcpp Sugar (e.g. `Rcpp::min()`).
    1. `all()`.
    2. `cumprod()`, `cummin()`, `cummax()`.
    3. `diff()`. Start by assuming lag 1, and then generalize for lag n.
    4. `range()`.
    
    - `all_cpp()` output examples:
    
        ```{Rcpp, eval = FALSE, echo = FALSE}
        // [[Rcpp::export]]
        bool all_cpp(LogicalVector x) {
          int n = x.length();
          for (int i = 0; i < n; i++) {
            if (!x[i]) {
              return false;
            }
          }
          return true;
        }
        ```
        
        <mark style="background-color: lightblue">R</mark>
            
        ```{r}
        all_cpp(c(TRUE, TRUE, FALSE))
        all_cpp(TRUE)
        all_cpp(FALSE)
        all_cpp(c(TRUE, TRUE))
        ```

    - `cumprod_cpp()` output examples
    
        ```{Rcpp, eval = FALSE, echo = FALSE}
        // [[Rcpp::export]]
        NumericVector cumprod_cpp(NumericVector x) {
          int n = x.length();
          if (n == 0 || n == 1) {
            return x;
          }
          
          NumericVector out(n);
          out[0] = x[0];
    
          for (int i = 1; i < n; i++) {
            out[i] = out[i - 1] * x[i];
          }
          
          return out;
        }
        ```
        
        <mark style="background-color: lightblue">R</mark>
            
        ```{r}
        x <- 1:4
        cumprod_cpp(x)
        x <- double()
        cumprod_cpp(x)
        x <- 3
        cumprod_cpp(x)
        ```
       
    - `cummin_cpp()` output examples:

        ```{Rcpp, eval = FALSE, echo = FALSE}
        // [[Rcpp::export]]
        NumericVector cummin_cpp(NumericVector x) {
          int n = x.length();
          if (n == 0 || n == 1) {
            return x;
          }
          
          NumericVector out(n);
          out[0] = x[0];
          
          for (int i = 1; i < n; i++) {
            out[i] = std::min(out[i - 1], x[i]);
          }
          
          return out;
        }
        ```
        
        <mark style="background-color: lightblue">R</mark>
                
        ```{r}
        cummin_cpp(double())
        cummin_cpp(3)
        cummin_cpp(c(10, 11, 3, 4, 5, 1, 2))
        ```

    - `cummax_cpp()` output examples:

        ```{Rcpp, eval = FALSE, echo = FALSE}
        // [[Rcpp::export]]
        NumericVector cummax_cpp(NumericVector x) {
          int n = x.length();
          if (n == 0 || n == 1) {
            return x;
          }
          
          NumericVector out(n);
          out[0] = x[0];
          
          for (int i = 1; i < n; i++) {
            out[i] = std::max(out[i - 1], x[i]);
          }
          
          return out;
        }
        ```
        
        <mark style="background-color: lightblue">R</mark>
                
        ```{r}
        cummax_cpp(double())
        cummax_cpp(3)
        cummax_cpp(c(10, 11, 3, 4, 5, 22, 2))
        ```

    - `diff_cpp()` output examples:

        ```{Rcpp, eval = FALSE, echo = FALSE}
        // [[Rcpp::export]]
        NumericVector diff_cpp(NumericVector x, int lag = 1) {
          int n = x.length();
          if (lag >= n) {
            NumericVector out(0);
            return out;
          }
          
          NumericVector out(n - lag);
          for (int i = 0; i < (n - lag); i++) {
            out[i] = x[i + lag] - x[i];
          }
          return out;
        }
        ```
        
        <mark style="background-color: lightblue">R</mark>
        
        ```{r}
        diff_cpp(x = c(1, 5, 10, 20), lag = 1)
        diff_cpp(x = c(1, 5, 10, 20), lag = 2)
        diff_cpp(x = c(1, 5, 10, 20), lag = 3)
        diff_cpp(x = c(1, 5, 10, 20), lag = 4)
        ```
    - `range_cpp()` output examples:
    
        ```{Rcpp, eval = FALSE, echo = FALSE}
        // [[Rcpp::export]]
        NumericVector range_cpp(NumericVector x) {
          int n = x.length();
          NumericVector out(2);
          if (n == 0) {
            out[0] = R_NegInf;
            out[1] = R_PosInf;
            return out;
          } else if (n == 1) {
            out[0] = x[0];
            out[1] = x[0];
            return out;
          }
        
          out[0] = x[0];
          out[1] = x[1];
          
          // do for-loop to get max and min, since avoiding sugar
          for (int i = 0; i < n; i++) {
            out[0] = std::min(out[0], x[i]);
            out[1] = std::max(out[1], x[i]);
          }
          
          return out;
        }
        ```
        
        <mark style="background-color: lightblue">R</mark>
        
        ```{r}
        range_cpp(c(9, 11, 22, 12, 1))
        range_cpp(8)
        range_cpp(double())
        ```
