---
title: "Armadillo"
author: "David Gerard"
date: "`r Sys.Date()`"
output:  
  html_document:
    toc: true
    toc_depth: 4
    toc_float: false
    highlight: pygments
urlcolor: "blue"
bibliography: "08_bib.bib"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo       = TRUE, 
                      fig.align  = "center",
                      fig.height = 3, fig.width = 4)
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(strip.background = ggplot2::element_rect(fill = "white")))
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy()
```

```{Rcpp, ref.label=knitr::all_rcpp_labels(), include=FALSE}
```

```{Rcpp, eval = FALSE, echo = FALSE}
#include <RcppArmadillo.h>
// [[Rcpp::depends(RcppArmadillo)]]
using namespace Rcpp;
using namespace arma;
```

# Learning Objectives

- [Armadillo Docs](http://arma.sourceforge.net/docs.html)
- Learning Objectives:
    - Using Armadillo for linear algebra
    
# Integration with R

- If you need to use numeric matrices or arrays in C++, you should use the Armadillo C++ library.

- [Eigen](https://cran.r-project.org/web/packages/RcppEigen/vignettes/RcppEigen-Introduction.pdf) is also a really great option, but appears to be less popular in the R community. You can read the docs [here](https://eigen.tuxfamily.org/dox/group__TutorialMatrixClass.html).

- Armadillo contains fast linear algebra algorithms, as well as a nice interface for working with matrices.

- You can set up an R package to use Armadillo by running the following R:

    <mark style="background-color: lightblue">R</mark>
    
    ```{r, eval = FALSE}
    usethis::use_rcpp_armadillo()
    ```
    
- This will add `RcppArmadillo` to the `LinkingTo` field in the `DESCRIPTION` file. It will also create `Makevars` files to set compilation settings.

- At the top of each C++ file, you now include the following (instead of just including Rcpp):

    <mark style="background-color: LightCoral">C++</mark>
    
    ``` cpp
    #include <RcppArmadillo.h>
    // [[Rcpp::depends(RcppArmadillo)]]
    using namespace Rcpp;
    using namespace arma;
    ```

# Armadillo Vectors

- If you are using Armadillo for matrices, you should be using Armadillo vectors instead of `NumericVector`s, so that it is easier to interact between matrices and vectors.

- Armadillo vectors are just like `NumericVector`'s with some minor differences. That is, there are no character or logical Armadillo vectors.

- Armadillo vectors are called `Col` objects (for "column vector").

- Initialize with

    <mark style="background-color: LightCoral">C++</mark>

    ``` cpp
    arma::vec x(n_elem)
    ```

- You can specify a default value for the vector with

    <mark style="background-color: LightCoral">C++</mark>

    ``` cpp
    // fill with zeros
    arma::vec x(n_elem, arma::fill::zeros);
    
    // fill with ones
    arma::vec y(3, arma::fill::ones);
    
    // fill with specified value
    arma::vec z(3, arma::fill::value(23.0));
    ```
    
- You can create custom vectors with braces

    <mark style="background-color: LightCoral">C++</mark>

    ``` cpp
    arma::vec w = {4.3, 2.2, -9.4};
    ```

- Let's demonstrate:

    ```{Rcpp, eval = FALSE}
    // [[Rcpp::export]]
    void ex1() {
      vec x(3, fill::zeros);
      Rcpp::Rcout << x << std::endl;
    
      vec y(3, fill::ones);
      Rcpp::Rcout << y << std::endl;
    
      vec z(3, fill::value(23.0));
      Rcpp::Rcout << z << std::endl;
      
      vec w = {4.3, 2.2, -9.4};
      Rcpp::Rcout << w << std::endl;
    }
    ```
    
    <mark style="background-color: lightblue">R</mark>
    
    ```{r}
    ex1()
    ```

- Of course, you typically get Armadillo vectors from a user.

    <mark style="background-color: LightCoral">C++</mark>

    ```{Rcpp, eval = FALSE}
    // [[Rcpp::export]]
    vec ex2(vec x) {
      return x;
    }
    ```
    
    <mark style="background-color: lightblue">R</mark>
    
    ```{r}
    ex2(c(1, 4, 2))
    ```
    
- If you return an Armadillo vector, then R will interpret this as a `matrix` object with one column. But you can pass R vectors as inputs to `Col` objects.
    
- You can get and set elements using parentheses notation `()`. You can use brackets as well, but this is more dangerous since Armadillo then won't do a bounds check. So you should always get and set elements by parentheses.

    <mark style="background-color: LightCoral">C++</mark>

    ```{Rcpp, eval = FALSE}
    // [[Rcpp::export]]
    void ex3() {
      vec x = {6.2, 5.3, 1.1};
      Rcpp::Rcout << "x(0): " << x(0) << std::endl;
      double y = x(1);
      Rcpp::Rcout << "y: " << y << std::endl;
      x(2) = 8.9;
      Rcpp::Rcout << "x(2): " << x(2) << std::endl;
    }
    ```
    
    <mark style="background-color: lightblue">R</mark>
    
    ```{r}
    ex3()
    ```

- You get the number elements in a vector with

    <mark style="background-color: LightCoral">C++</mark>

    ``` cpp
    x.n_elem
    ```

- You do arithmetic operations on vectors as you would in R, but note that element-wise multiplication is different.

    <mark style="background-color: LightCoral">C++</mark>

    ``` cpp
    x + y
    x - y
    x % y // element-wise multiplication
    x / y
    ```

- You can do lots of miscellaneous element-wise functions (table from [Armadillo docs](http://arma.sourceforge.net/docs.html#operators)):

    | Armadillo |        Explanation    |
    |-----------|-----------------------|
    | `exp(A)` | base-e exponential: $e^x$ |
    | `exp2(A)` | base-2 exponential: $2^x$ |
    | `exp10(A)` | base-10 exponential: $10^x$ |
    | `expm1(A)` | compute $\exp(A)-1$ accurately for values of A close to zero |
    | `log(A)` | natural log: $\log_e(x)$ |
    | `log2(A)` | base-2 log: $\log_2(x)$ |
    | `log10(A)` | base-10 log: $\log_{10}(x)$ |
    | `log1p(A)` | compute $\log(1+A)$ accurately for values of A close to zero |
    | `pow(A, p)` | raise to the power of p: $x^p$ |
    | `square(A)` | square: $x^2$ |
    | `sqrt(A)` | square root: $\sqrt{x}$ |
    | `floor(A)` | largest integral value that is not greater than the input value |
    | `ceil(A)` | smallest integral value that is not less than the input value |
    | `round(A)` | round to nearest integer, with halfway cases rounded away from zero |
    | `trunc(A)` | round to nearest integer, towards zero |
    | `lgamma(A)` | natural log of the absolute value of gamma function |
    | `sign(A)` | sign function |


- The usual statistics functions are available, but note the different name for standard deviation: `mean()`, `median()`, `stddev()`, `var()`, `range()`

- **Exercise**: Implement `cumprod()`, `cummin()`, `cummax()`, and `diff()` using Armadillo vectors instead of `NumericVector`s. You can copy and paste your solutions from the exercises in the [Rcpp vectors lecture](./08_cpp_vectors.html) and then modify them to use armadillo vectors. 

    *Hint*: You can make a length zero `Col` object by
    
    <mark style="background-color: LightCoral">C++</mark>
    
    ``` cpp
    vec x(1);
    x.clear();
    ```
    
    ```{Rcpp, eval = FALSE, echo = FALSE}
    // [[Rcpp::export]]
    vec cumprod_arma(vec x) {
      int n = x.n_elem;
      if (n == 0 || n == 1) {
        return x;
      }
      
      vec out(n);
      out(0) = x(0);

      for (int i = 1; i < n; i++) {
        out(i) = out(i - 1) * x(i);
      }
      
      return out;
    }
    ```
        
    ```{r, echo = FALSE, eval = FALSE}
    x <- 1:4
    cumprod_arma(x)
    x <- double()
    cumprod_arma(x)
    x <- 3
    cumprod_arma(x)
    ```

    ```{Rcpp, eval = FALSE, echo = FALSE}
    // [[Rcpp::export]]
    vec cummin_arma(vec x) {
      int n = x.n_elem;
      if (n == 0 || n == 1) {
        return x;
      }
      
      vec out(n);
      out(0) = x(0);
      
      for (int i = 1; i < n; i++) {
        out(i) = std::min(out(i - 1), x(i));
      }
      
      return out;
    }
    ```
            
    ```{r, echo = FALSE, eval = FALSE}
    cummin_arma(double())
    cummin_arma(3)
    cummin_arma(c(10, 11, 3, 4, 5, 1, 2))
    ```

    ```{Rcpp, eval = FALSE, echo = FALSE}
    // [[Rcpp::export]]
    vec cummax_arma(vec x) {
      int n = x.n_elem;
      if (n == 0 || n == 1) {
        return x;
      }
      
      vec out(n);
      out(0) = x(0);
      
      for (int i = 1; i < n; i++) {
        out(i) = std::max(out(i - 1), x(i));
      }
      
      return out;
    }
    ```
            
    ```{r, echo = FALSE, eval = FALSE}
    cummax_arma(double())
    cummax_arma(3)
    cummax_arma(c(10, 11, 3, 4, 5, 22, 2))
    ```

    ```{Rcpp, eval = FALSE, echo = FALSE}
    // [[Rcpp::export]]
    vec diff_arma(vec x, int lag = 1) {
      int n = x.n_elem;
      if (lag >= n) {
        vec out(1);
        out.clear();
        return out;
      }
      
      vec out(n - lag);
      for (int i = 0; i < (n - lag); i++) {
        out(i) = x(i + lag) - x(i);
      }
      return out;
    }
    ```
    
    ```{r, echo = FALSE, eval = FALSE}
    diff_arma(x = c(1, 5, 10, 20), lag = 1)
    diff_arma(x = c(1, 5, 10, 20), lag = 2)
    diff_arma(x = c(1, 5, 10, 20), lag = 3)
    diff_arma(x = c(1, 5, 10, 20), lag = 4)
    ```

# Armadillo Matrices

- Armadillo calls matrix objects `Mat`. They contain doubles.

- You initialize Armadillo matrices with

    <mark style="background-color: LightCoral">C++</mark>

    ``` cpp
    arma::max A(n_rows, n_cols);
    ```

- You can specify a default value for the vector with

    <mark style="background-color: LightCoral">C++</mark>

    ``` cpp
    // fill with zeros
    arma::mat A1(n_rows, n_cols, arma::fill::zeros);
    
    // fill with ones
    arma::mat A2(n_rows, n_cols, arma::fill::ones);
    
    // fill with specified value
    arma::mat A3(n_rows, n_cols, arma::fill::value(23.0));
    ```
    
- You can create custom matrices with braces. Internal braces are the rows.

    <mark style="background-color: LightCoral">C++</mark>

    ``` cpp
    // Three rows, two columns
    arma::mat A4 = {{4.3, 2.2}, {-9.4, 7.0}, {10.0, 11.0}};
    ```

- Let's demonstrate these    

    <mark style="background-color: LightCoral">C++</mark>

    ```{Rcpp, eval = FALSE}
    // [[Rcpp::export]]
    void ex4() {
      int n_rows = 3;
      int n_cols = 2;
      
      arma::mat A1(n_rows, n_cols, arma::fill::zeros);
      Rcpp::Rcout << A1 << std::endl;
    
      arma::mat A2(n_rows, n_cols, arma::fill::ones);
      Rcpp::Rcout << A2 << std::endl;
      
      arma::mat A3(n_rows, n_cols, arma::fill::value(23.0));
      Rcpp::Rcout << A3 << std::endl;
      
      arma::mat A4 = {{4.3, 2.2}, {-9.4, 7.0}, {10.0, 11.0}};
      Rcpp::Rcout << A4 << std::endl;
    }
    ```
    
    <mark style="background-color: lightblue">R</mark>
    
    ```{r}
    ex4()
    ```

- Element-wise arithmetic operations are the same as for vectors:

    <mark style="background-color: LightCoral">C++</mark>

    ``` cpp
    A + B;
    A - B;
    A % B; // element-wise multiplication
    A / B;
    ```
    
- Element-wise functions (e.g. `log()` and `exp()`) are the exact same as in vectors.

- You get and set elements by parentheses notation. Always remember that **indexing starts at 0**.

    <mark style="background-color: LightCoral">C++</mark>

    ``` cpp
    A(0, 2); // get (0,2)th element
    double x = A(4, 2); // get (4,2)th element and assign to x
    A(1, 0) = 12.0; // set (1,0)th element
    ```

- Let's demonstrate these

    <mark style="background-color: LightCoral">C++</mark>

    ```{Rcpp, eval = FALSE}
    // [[Rcpp::export]]
    void ex5() {
      mat A = {{1.0, 2.0, 3.0}, {4.0, 5.0, 6.0}};
      
      Rcpp::Rcout << "A: " << A << std::endl;
      
      Rcpp::Rcout << "A(0, 2): " << A(0, 2) << std::endl;
      
      A(1, 0) = 12.0;
      Rcpp::Rcout << "A(1, 0): " << A(1, 0) << std::endl;
      
      A = A + 2.0;
      Rcpp::Rcout << "A: " << A << std::endl;
    }
    ```
    
    <mark style="background-color: lightblue">R</mark>
    
    ```{r}
    ex5()
    ```

- Useful fields for `Mat` fields:

    - `A.n_elem`: Number of elements in matrix `A`.
    - `A.n_rows`: Number of rows in matrix `A`.
    - `A.n_cols`: Number of columns in matrix `A`.
    
- Generate matrix of ones.

    ``` cpp
    mat A = ones<mat>(n_rows, n_cols);
    ```

- Generate $n\times n$ identity matrix.

    ``` cpp
    mat id = eye(n_rows, n_cols);
    ```

- Diagonal matrix, where `v` is a `Col` that you already have defined.

    ``` cpp
    mat d = diagmat(v);
    ```
    
- Let's demonstrate making these special matrices:
    
    <mark style="background-color: LightCoral">C++</mark>
    
    ```{Rcpp, eval = FALSE}
    // [[Rcpp::export]]
    void ex6() {
      mat A = ones<mat>(2, 3);
      Rcpp::Rcout << "A: " << std::endl << A << std::endl;
      
      mat id = eye(4, 4);
      Rcpp::Rcout << "id: " << std::endl << id << std::endl;
      
      vec v = {3.0, 2.2, 5.5};
      mat d = diagmat(v);
      Rcpp::Rcout << "d: " << std::endl << d << std::endl;
    }
    ```
    
    <mark style="background-color: lightblue">R</mark>
        
    ```{r}
    ex6()
    ```

# Linear Algebra Operations

- Rcpp has `LogicalMatrix`, `IntegerMatrix`, `NumericMatrix`, and `CharacterMatrix` objects that I am not covering.

- This is because you mostly use C++ for numerical computing, so it's less usual to use a `CharacterMatrix` in C++ (I would say you are crazy for trying to use C++ when you are mostly dealing with strings). And for numerical computing, linear algebra is king, and Armadillo has nice linear algebra routines.

- [Matrix multiplication](https://en.wikipedia.org/wiki/Matrix_multiplication):

    <mark style="background-color: LightCoral">C++</mark>

    ``` cpp
    A * B;
    ```

- [Matrix transpose](https://en.wikipedia.org/wiki/Transpose):

    <mark style="background-color: LightCoral">C++</mark>

    ``` cpp
    A.t();
    ```

- [Solving linear systems](https://en.wikipedia.org/wiki/System_of_linear_equations#Solving_a_linear_system): Solving for $X$ in
    $$
    AX = B
    $$
    results in
    $$
    X = A^{-1}B
    $$
    `solve()` will calculate $A^{-1}B$ numerically stably.
    
    <mark style="background-color: LightCoral">C++</mark>
    
    ``` cpp
    solve(A, B);
    ```
    
- [Matrix inverse](https://en.wikipedia.org/wiki/Moore%E2%80%93Penrose_inverse):

    <mark style="background-color: LightCoral">C++</mark>

    ``` cpp
    inv(A);
    ```

- [Moore-Penrose inverse](https://en.wikipedia.org/wiki/Moore%E2%80%93Penrose_inverse):

    <mark style="background-color: LightCoral">C++</mark>

    ``` cpp
    pinv(A);
    ```

- [Extracting diagonal](https://en.wikipedia.org/wiki/Main_diagonal):

    <mark style="background-color: LightCoral">C++</mark>

    ``` cpp
    A.diag();
    ```

- [Matrix trace](https://en.wikipedia.org/wiki/Trace_(linear_algebra)):

    <mark style="background-color: LightCoral">C++</mark>

    ``` cpp
    trace(A);
    ```

- [Cholesky Decomposition](https://en.wikipedia.org/wiki/Cholesky_decomposition): Find an [upper triangular](https://en.wikipedia.org/wiki/Triangular_matrix) matrix $R$ with non-negative diagonal elements such that $X = R^TR$. Applicable when $X$ is [symmetric](https://en.wikipedia.org/wiki/Symmetric_matrix).

    <mark style="background-color: LightCoral">C++</mark>

    ``` cpp
    mat R = chol(X);
    ```

- [Eigendecomposition](https://en.wikipedia.org/wiki/Eigendecomposition_of_a_matrix): Find [orthogonal matrix](https://en.wikipedia.org/wiki/Orthogonal_matrix) $U$ and [diagonal matrix](https://en.wikipedia.org/wiki/Diagonal_matrix) $D$ such that $X = UDU^T$. Applicable when $X$ is [symmetric](https://en.wikipedia.org/wiki/Symmetric_matrix).

    You need to create (i) a vector containing the main diagonal of $D$ and (ii) a matrix that will hold the eigenvectors. You do that ahead of time. Then you run `eig_sym()`.
    
    <mark style="background-color: LightCoral">C++</mark>

    ``` cpp
    vec eigval; // main diagonal of D matrix
    mat eigvec; // U matrix
    eig_sym(eigval, eigvec, B);
    ```

- [Singular Value Decomposition](https://en.wikipedia.org/wiki/Singular_value_decomposition): Find orthogonal matrices $U$ and $V$, and diagonal matrix $D$ such that $X = UDV^T$.

    You need to create (i) a vector containing the main diagonal of $D$ and (ii) matrices that will hold the left and right singular vectors. You do that ahead of time. Then you run `svd()`.
    
    <mark style="background-color: LightCoral">C++</mark>

    ``` cpp
    mat U;
    vec s;
    mat V;
    svd(U,s,V,X);
    ```
    
- [QR Decomposition](https://en.wikipedia.org/wiki/QR_decomposition): Find an orthogonal matrix $Q$ and an upper triangular matrix $R$ with non-negative diagonal entries, such that $X = QR$.

    <mark style="background-color: LightCoral">C++</mark>

    ``` cpp
    mat Q;
    mat R;
    qr(Q, R, X);
    ```

- **Example**: This isn't the most efficient way to find the OLS estimates from a regression, but we can just calculate

    $$
    \hat{\beta} = (X^TX)^{-1}X^Ty
    $$

    Let's do this in C++:
    
    <mark style="background-color: LightCoral">C++</mark>
    
    ```{Rcpp, eval = FALSE}
    // [[Rcpp::export]]
    vec stupid_ols(mat X, vec y) {
      vec betahat = inv(X.t() * X) * X.t() * y;
      return betahat;
    }
    ```
    
    Let's verify our results.
    
    <mark style="background-color: lightblue">R</mark>
    
    ```{r}
    ## Armadillo Way
    X <- model.matrix(mpg ~ wt, data = mtcars)
    y <- mtcars$mpg
    stupid_ols(X = X, y = y)
    
    ## R Way
    coef(lm(mpg ~ wt, data = mtcars))
    ```
    
- I called it `stupid_ols()` because calculating an inverse is generally a computationally unstable task (but in this scenario, it's probably OK). It's better to use `solve()` here, but folks have thought very hard about this simple problem [@farebrother1988linear]:

    <mark style="background-color: LightCoral">C++</mark>

    ```{Rcpp, eval = FALSE}
    // [[Rcpp::export]]
    vec arma_ols(mat X, vec y) {
      vec betahat = solve(X.t() * X, X.t() * y);
      return betahat;
    }
    ```

    <mark style="background-color: lightblue">R</mark>
      
    ```{r}
    arma_ols(X = X, y = y)
    ```
    
- Let's compare performance when we have a large number of covariates.

    <mark style="background-color: lightblue">R</mark>
    
    ```{r}
    X <- model.matrix(mpg ~ ., data = mtcars)
    y <- mtcars$mpg
    
    bench::mark(
      coef(lm(mpg ~ wt, data = mtcars)),
      stupid_ols(X = X, y = y),
      arma_ols(X = X, y = y), 
      check = FALSE
    )[1:8] |>
      knitr::kable()
    ```


# R to Armadillo Conversion Table

In the below table, `A` and `B` are a matrices and `Q` is a three-dimensional array (or "cube"). This table is based on the Matlab/Octave to Armadillo conversion table [here](http://arma.sourceforge.net/docs.html#syntax).

| R | Armadillo | Notes |
|---|---|-----|
| `A[1, 1]` | `A(0, 0)` | indexing in Armadillo starts at 0 | 
| `A[k, k]` | `A(k-1, k-1)` |  | 
| `nrow(A)` | `A.n_rows` | read only | 
| `ncol(A)` | `A.n_cols` |  | 
| `dim(A)[[3]]` | `Q.n_slices` | `Q` is a cube (3D array) | 
| `length(A)` | `A.n_elem` |  | 
| `A[, k]` | `A.col(k)` | this is a conceptual example only; exact conversion from R to Armadillo syntax will require taking into account that indexing starts at 0 | 
| `A[k, ]` | `A.row(k)` |  | 
| `A[, p:q]` | `A.cols(p,q)` |  | 
| `A[p:q, ]` | `A.rows(p,q)` |  | 
| `A[p:q, r:s]` | `A(span(p,q),span(r,s))` | A(span(first_row,last_row), span(first_col,last_col)) | 
| `Q[, , k]` | `Q.slice(k)` | `Q` is a cube (3D array) | 
| `Q[, , t:u]` | `Q.slices(t,u)` |  | 
| `Q[p:q, r:s, t:u]` | `Q(span(p,q),span(r,s),span(t,u))` |  | 
| `t(A)` | `A.t()`or`trans(A)` | matrix transpose |
| `A[] <- 0` | `A.zeros()` | Fill a matrix with 0's | 
| `A[] <- 1` | `A.ones()` | Fill a matrix with 1's | 
| `A <- matrix(0, nrow = k, ncol = k)` | `A = zeros<mat>(k,k)` | Initialize a 0 matrix | 
| `A <- matrix(1, nrow = k, ncol = k)` | `A = ones<mat>(k,k)` | Initialize a 1 matrix | 
| `A * B` | `A % B` | element-wise multiplication | 
| `A / B` | `A / B` | element-wise division | 
| `solve(A, B)` | `solve(A, B)` | Solve linear equation $Ax = B$ for $A$ | 
| `A <- A + 1;` | `A++` |  | 
| `A <- A - 1;` | `A--` |  | 
| `A <- matrix(c(1, 2, 3, 4), nrow = 2, ncol = 2, byrow = TRUE)` | `A = {{1,2}, {3,4}}` | element initialization | 
| `X <- c(A)` | `X = vectorise(A)` |  | 
| `X = cbind(A, B)` | `X = join_horiz(A, B)` |  | 
| `X = rbind(A, B)` | `X = join_vert(A, B)` |  | 
| `A` | `cout << A << endl` or `A.print("A=")` |  | 

# References
