---
title: "S3 Objects"
author: "David Gerard"
date: "`r Sys.Date()`"
output:  
  html_document:
    toc: true
    toc_depth: 4
    toc_float: false
    highlight: pygments
urlcolor: "blue"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo       = TRUE, 
                      fig.align  = "center",
                      fig.height = 3, fig.width = 4)
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(strip.background = ggplot2::element_rect(fill = "white")))
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy()
```

# Learning Objectives

- Chapter 13 of [Advanced R](https://adv-r.hadley.nz/)
- [`{vctrs}`](https://vctrs.r-lib.org/articles/s3-vector.html) package

# Motivation

- S3 is the most commonly used object-oriented programming (OOP) system in R.

- Most of the common data types you are used to are S3.

    ```{r}
    # Data frames are S3
    sloop::otype(mtcars)
    
    # lm objects are S3
    lmout <- lm(mpg ~ wt, data = mtcars)
    sloop::otype(lmout) 
    
    # ggplot2 plots are S3
    pl <- ggplot2::ggplot(mtcars, ggplot2::aes(x = wt, y = mpg)) + 
      ggplot2::geom_point()
    sloop::otype(pl)
    
    # tidymodels use S3
    tdout <- 
      parsnip::linear_reg() |>
      parsnip::set_engine("lm") |>
      parsnip::fit(mpg ~ wt, data = mtcars)
    sloop::otype(tdout)
    
    # Factors are S3
    x <- factor(c(1, 2, 3))
    sloop::otype(x)
    
    # Dates are S3
    x <- lubridate::make_date(year = 1970, month = 1, day = 1)
    sloop::otype(x)
    ```

- If you are creating a package and you want OOP features, you should use S3 unless
    1. You work in a large team (use S4 then).
    2. Modify-by-reference is important (use R6).
    
- This is since most R programmers are used to S3 (intuitively) and are not used to S4 or R6.

# S3 Basics

- An S3 object is any variable with a `class` attribute. This is the full definition.

- S3 objects may or may not have more attributes.

- E.g. the `factor` class always has the `levels` attribute.

    ```{r}
    x <- factor(c("A", "B", "B", "A", "C", "A"))
    attributes(x)
    ```

- You can get the underlying bayes type by `unclass()`.

    ```{r}
    unclass(x)
    ```

- S3 objects behave differently when passed to a **generic** function.

- Use `sloop::ftype()` to see if a function is generic.

- These are all generics
    ```{r}
    sloop::ftype(print)
    sloop::ftype(summary)
    sloop::ftype(plot)
    ```

- But these are not:
    ```{r}
    sloop::ftype(sum)
    sloop::ftype(c)
    ```

# New functions

- `unclass()`: Remove class attribute and obtain underlying base type.
- `sloop::ftype()`: See if a function is a "regular/primitive/internal function, a internal/S3/S4 generic, or a S3/S4/RC method".
