---
title: "S4 Objects"
author: "David Gerard"
date: "`r Sys.Date()`"
output:  
  html_document:
    toc: true
    toc_depth: 4
    toc_float: false
    highlight: pygments
urlcolor: "blue"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo       = TRUE, 
                      fig.align  = "center",
                      fig.height = 3, fig.width = 4)
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(strip.background = ggplot2::element_rect(fill = "white")))
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy()
```

# Learning Objectives

- Chapter 15 of [Advanced R](https://adv-r.hadley.nz/)
- [Converting Packages to S4](https://cran.r-project.org/doc/Rnews/Rnews_2003-1.pdf)
- [Introduction to Matrix Package](https://cran.r-project.org/web/packages/Matrix/vignettes/Intro2Matrix.pdf)

# Motivation

- Because S3 is really informal, it is also really easy to make dangerous mistakes.

- **Example**: Class structures are not enforced.
    - E.g. you can do 
        ```{r}
        mt <- mtcars
        class(mt) <- "lm"
        ```
        and method dispatch will try to use `lm` methods on `mt` even though that makes no sense. 
    - This is because in S3 the class *is* the `class` attribute.
    - The structure of a class is not enforced, except by perhaps a "gentleman's agreement" that classes should be of a certain form.

- **Example**: Using function names to identify methods is dangerous. E.g. is `row.names.data.frame()`
    1. The `row()` method of the `names.data.frame` class?
    2. The `row.names()` method of the `data.frame` class?
    3. The `row.names.data()` method of the `frame` class?

- S3 is also limiting because the method is only determined by the class of one object. But what if you have a function `f(x, y)` and you want the method to be different based on:
    - `x` and `y` are both class `foo`.
    - `x` is class `foo` and `y` is class `bar`.
    - `x` is class `bar` and `y` is class `foo`.
    - `x` and `y` are both class `bar`
    
- S4 
    1. Makes creating objects of a class more strict, to enforce structure.
    2. Makes method creation more strict, so you don't have to name a new function `generic.class()` and just hope for the best.
    3. Allows for multiple dispatch, where you can choose the method based on the class of more than one object.

# `{methods}` Package

- All functions for S4 are in the `{methods}` package which comes with R by default.
    - So in a package you either need to import the `{methods}` package, or you need to use `::` to use these functions create S4 objects.
    - Running R in batch mode (e.g. running it on the supercomputer) also does not attach the `{methods}` package, so you need to do `library(methods)` if you intend to use S4 in batch mode.

# S4 Classes

- In S3, *any* object with a `class` attribute is an S3 object.

- In S4, you need to explicitly run the `methods::setClass()` function to define the class. This makes the structure explicitly enforced.
    - The `Class` argument is the name of the new class.
    - The `slots` argument contains a *named* character vector with the name of the fields in the class and the type for each field. 
        - Note that it supports `"numeric"` and not `"double"`.
    - The `prototype` argument contains the default values for each slot. This is typically either
        1. A names list object, or
        2. If inheriting from a basetype vector, you use `structure`, where the base type is in an implicit slot called `.Data` and the rest of the elements are the other slots. See below for example usage.

    ```{r}
    setClass(
      Class = "Person",
      slots = c(
        name = "character",
         age = "numeric"
      ),
      prototype = list(
        name = character(),
        age = double()
      )
    )
    ```

- *Note*: In S4, they call fields **slots** (i.e. data that belongs to the class).

- You then use `methods::new()` to create objects of the given S4 class, specifying the class and filling in the slots.

    ```{r}
    david <- new(Class = "Person", name = "David Gerard", age = 34)
    ```

- `david` is an S4 object
    ```{r}
    sloop::otype(david)
    ```

- You use `methods::is()` to see (via one argument) and test (via two arguments) the S4 class of an object.

    ```{r}
    is(david)
    is(david, "Person")
    ```

- You extract the data from a slot with `methods::slot()`

    ```{r}
    slot(david, "age")
    ```

- Equivalently, you can use `@` notation to extract slots. (You do not use `$` or `[[` notation for S4 objects).

    ```{r}
    david@age
    ```
    
    ```{r, error = TRUE}
    ## will error
    david$age
    ## will error
    david[["age"]]
    ```

- The `@` notation is typically only used *internally.* If you want folks to access slots, you need to define an **accessor** function. Accessor's are usually methods for S4 generics, which we will cover later.

## Inheritance

- Inheritance is really informal in S3. `class` is a vector and method dispatch just sequentially goes down the vector. But there is no guarantee that a subclass will have the correct structure of a superclass, or that the `class` vector is consistent across all members of a class.

- In S4, use the `contains` argument in `setClass()` to explicitly define the superclass that your class is inheriting from.

- E.g. we can create a class called `Employee` that inherits from `Person`.

    ```{r}
    setClass(
      Class = "Employee", 
      contains = "Person",
      slots = c(
        boss = "character",
        years = "numeric"
      ),
      prototype = list(
        boss = NA_character_,
        years = NA_real_
      )
    )
    ```

- Now, we can create an `Employee` object with all of the slots of `Person`, along with the new slots from `Employee`.

    ```{r}
    david <- new(Class = "Employee", 
                 name = "David Gerard", 
                 age = 34, 
                 boss = "Self", 
                 years = 4)
    david
    ```

## Helper Function

- You should not expect users to use `new()`. Instead you should create a helper function for them. E.g.

    ```{r}
    Employee <- function(name, 
                         age = NA_real_, 
                         boss = NA_character_, 
                         years = NA_real_) {
      age <- as.double(age)
      boss <- as.character(boss)
      years <- as.double(years)
      
      return(new("Employee", name = name, age = age, boss = boss, years = years))
    }
    ```
    
    ```{r}
    david <- Employee(name = "David Gerard", age = 34, boss = "Self", years = 4)
    david
    ```

- Try to make sensible defaults, and try to type-cast as much as you can.

## Validator Functions

- Suppose we want `Person` to be a vector-type object. So we should constrain the `age` and `name` slots to be the same size.

- This is not currently a constraint
    ```{r}
    profs <- new(Class = "Person", name = c("David Gerard", "Zois Boukouvalas"), age = 34)
    profs
    ```

- You define S4 validators with `setValidity()`.
    - `Class` is the name of the class to create a validator for.
    - `method` is the function that should be run on the class.
        - It returns `TRUE` if all validity checks pass.
        - It returns a character string describing the issues if some of the validity checks fail.

    ```{r}
    setValidity(
      Class = "Person", 
      method = function(object) {
        if(length(object@name) != length(object@age)) {
          "@name and @age must be same length"
        } else {
          return(TRUE)
        }
      }
    )
    ```

- Now it is impossible to make `@name` and `@age` slots of different lengths
    ```{r, error = TRUE}
        profs <- new(Class = "Person", name = c("David Gerard", "Zois Boukouvalas"), age = 34)
    ```

- Note: Validity checks only run automatically with `new()`, so you can edit the slot to have invalid inputs.

    ```{r}
    profs <- new(Class = "Person",
                 name = c("David Gerard", "Zois Boukouvalas"),
                 age = c(34, NA_real_))
    profs@age <- 1:20
    ```

- But you can manually check validity with `validObject()`.

    ```{r, error = TRUE}
    validObject(profs)
    profs <- new(Class = "Person", 
                 name = c("David Gerard", "Zois Boukouvalas"), 
                 age = c(34, NA_real_))
    validObject(profs)
    ```

- **Exercise**: Create a `factor2` S4 class that tries to implement the `factor` class for S4.
    - Define the class (with appropriate slots and prototypes)
    - Create a helper function. It should be able to take a character vector as input and define the 
    - Create a validator. It should check that the the maximum integer is less than or equal to the number of levels.

    ```{r, eval = FALSE, echo = FALSE}
    setClass(
      Class = "factor2",
      slots = c(
        x = "integer",
        levels = "character"
      ),
      prototype = list(
        x = integer(),
        levels = character()
      )
    )
    
    setValidity(
      Class = "factor2", 
      method = function(object) {
        if (length(object@x) == 0 & length(object@levels) == 0) {
          return(TRUE)
        } else if (length(object@levels) < max(object@x)) {
          return(paste0("Number of levels must be greater than or equal to", max(object@x)))
        } else {
          return(TRUE)
        }
      }
    )
    
    factor2 <- function(x) {
      levels <- sort(unique(x))
      dat <- match(x, levels)
      return(new(Class = "factor2", x = dat, levels = levels))
    }
    
    factor2(c("A", "A", "B", "A", "B"))
    ```



# S4 Generics and Methods

- You create generic functions with `setGeneric()`.
    - The `name` is the name of the new generic.
    - The `def` is the function of the generic. Recall that S3 generics are really simple and just call `UseMethod("generic")`. Similarly, S4 generic functions should be really simple and just call `standardGeneric("generic")`.
    
- Below we are defining a generic called `age()` (for getting the age of individuals) and a generic for `age()<-` which will act as a replacement function.
    ```{r}
    setGeneric(name = "age", 
               def = function(x) standardGeneric("age"))
    setGeneric(name = "age<-", 
               def = function(x, value) standardGeneric("age<-"))
    ```

- Literally most S4 generics look like this
    ```{r}
    BiocGenerics::append
    BiocGenerics::dims
    BiocGenerics::get
    ```

- The only other argument you should usually use in `setGeneric()` is `signature`.
    - By default, method dispatch is determined by *all* of the arguments in the `def` function
    - You can choose to only do method dispatch based on some arguments via signature. 
    - E.g. below only `x` determines method dispatch:
        ```{r}
        setGeneric(name = "age<-", 
                   def = function(x, value) standardGeneric("age<-"),
                   signature = "x")
        ```

- You define a method for a given generic with `setMethod()`
    - `f`: The name of the generic.
    - `signature`: The name of the class(es) to which to apply the method.
    - `definition`: The method function.

    ```{r}
    setMethod(f = "age", 
              signature = "Person", 
              definition = function(x) {
                return(x@age)
              }
             )
    setMethod(f = "age<-",
              signature = "Person",
              definition = function(x, value) {
                x@age <- value
                return(x)
              }
             )
    ```

- Let's look at these

    ```{r}
    david <- Employee(name = "David Gerard", age = 34)
    age(david)
    ```

- Notice that this worked with an `Employee` object even though we only defined the method for `Person`. This is because `Employee` inherits from `Person`.

- Let's reassign age
    ```{r}
    age(david) <- 21 # ah, young again
    age(david)
    ```

- The `age()` method is called an **accessor** or, more informally, a **getter**.

- The `age()<-` method is called a **replacement function** or, more informally, a **setter**.

# Getting Help

- To read the help file of an S4 class, put `class?` before the name of the class. E.g. the following will open up the help file of the `GRanges` class from the `{GenomicRanges}` package from Bioconductor.

    ```{r}
    library(stats4)
    ```

    ```{r, eval = FALSE}
    class?mle
    ```
    
- This is equivalent to
    ```{r, eval = FALSE}
    ?`mle-class`
    ```
    
- You can see the code of a generic function with `getGeneric()`
    ```{r}
    getGeneric(f = "show")
    ```

- You can obtain the method of a generic with `getMethod()`
    ```{r}
    getMethod(f = "show", signature = "mle")
    ```
    
- You can use `args()` on both of these to get their arguments
    ```{r}
    args(getGeneric("show"))
    args(getMethod("show", "mle"))
    ```

- To read the help file of method, put a question mark before the call of a generic. E.g., to see the help file for the `coverage()` method of a `GRanges` object, run

    ```{r, eval = FALSE}
    ## Create mle object
    x <- c(1, 2, 3, 4)
    gobj <- mle(minuslogl = function(mu) -sum(dnorm(x = x, mean = mu, log = TRUE)), start = 0) 
    
    ## opens help file for show method of mle
    ?show(gobj) 
    ```

- `{sloop}` is useful for seeing 
    1. If an object is an S4 object
        ```{r}
        sloop::otype(david)
        ```
    2. If a function is an S4 generic
        ```{r}
        sloop::ftype(age)
        ```
    3. A list of all methods associated with an S4 class
        ```{r}
        sloop::s4_methods_class("Person")
        ```
    4. A list of all methods associated with an S4 generic
        ```{r, warning = FALSE}
        sloop::s4_methods_generic("age")
        ```

## `show()` method

- You should always define a `show()` method. This is used for printing S4 objects.

- You need to use the same arguments of a generic that already exists. E.g. show needs `object`
    ```{r}
    args(getGeneric("show"))
    ```

- You usually use the `cat()` function for printing.

    ```{r}
    setMethod(
      f = "show", 
      signature = "Person", 
      definition = function(object) {
        cat(is(object)[[1]], "\n",
            "  Name: ", object@name, "\n",
            "   Age: ", object@age, "\n",
            sep = "")
      }
    )
    ```

- Now we get better printing for `Person`'s and `Employee`'s.

    ```{r}
    david
    ```

# Accessors

- Above, we defined the accessor `age` as a generic. 

- But if you have slots unique to your class, you don't need to go all out OOP like that.

- A basic accessor can be defined as follows:

    ```{r}
    person_name <- function(x) x@name
    person_name(david)
    ```
    
- Or we can go through the whole rigmarole so that other classes can use your generic.

    ```{r}
    setGeneric(name = "name", def = function(x) standardGeneric("name"))
    setMethod(f = "name", signature = "Person", definition = function(x) x@name)
    
    setGeneric(name = "name<-", def = function(x, value) standardGeneric("name<-"))
    setMethod(
      f = "name<-", 
      signature = "Person",
      definition = function(x, value) {
        x@name <- value
        validObject(x)
        return(x)
      }
    )
    ```
    
    ```{r}
    name(david)
    name(david) <- "DG"
    david
    ```

- **Exercise**: Define a `show()` method for the `factor2` class. It should have similar output as
    ```{r, eval = FALSE, echo = FALSE}
    setMethod(
      f = "show",
      signature = "factor2",
      definition = function(object) {
        cat(object@levels[object@x], sep = " ")
      }
    )
    x <- factor2(c("A", "B", "A", "A", "B"))
    x
    ```


- **Exercise**: Define a `levels2()` and `levels2<-` generic and method for the `factor2` class that will set and get the `levels` slot.

    ```{r, eval = FALSE, echo = FALSE}
    setGeneric(name = "levels2", def = function(x) standardGeneric("levels2"))
    setGeneric(name = "levels2<-", def = function(x, value) standardGeneric("levels2<-"))
    
    setMethod(
      f = "levels2", 
      signature = "factor2", 
      definition = function(x) {
        return(x@levels)
      }
    )
    
    setMethod(
      f = "levels2<-", 
      signature = "factor2", 
      definition = function(x, value) {
        x@levels <- value
        return(x)
      }
    )
    
    
    levels2(x) <- c("C", "D")
    levels2(x)
    x
    
    ## Subsetting still works
    levels2(x)[1] <- "B"
    levels2(x)
    x
    ```

# Coercion

- Use `setAs()` to define a method for coercing an object to a given class.
    - `from`: The class you are coercing from.
    - `to`: The class you are coercing to.
    - `def` A function of one argument, of class `from` that will convert it to class `to`. Typically the name of the argument is just `from`.

- Let's coerce a `character` vector into a `Person` by assuming that the character vector contains the names.

    ```{r}
    setAs(
      from = "character",
      to = "Person",
      def = function(from) {
        return(
          new(Class = "Person", 
              name = from, 
              age = rep(NA_real_, length.out = length(from)))
        )
      }
    )
    ```

- You then use `methods::as()` to apply the coercion.

    ```{r}
    pvec <- as(object = c("David", "Alice", "John", "Jenny"), Class = "Person")
    pvec
    ```

- Though, just as with `new()` and `is()`, you try to hide `as()` from the user through a generic function

    ```{r}
    setGeneric(name = "asPerson", def = function(x) standardGeneric("asPerson"))
    
    setMethod(
      f = "asPerson", 
      signature = "character", 
      definition = function(x) {
        as(object = x, Class = "Person")
      }
    )
    
    pvec <- asPerson(c("David", "Bob", "Megan"))
    pvec
    ```


- **Exercise**: Create a coercion for `factor2` from `character`.
    ```{r, eval = FALSE, echo = FALSE}
    setAs(from = "character", 
          to = "factor2",
          def = function(from) {
            levels <- sort(unique(from@.Data))
            return(new(Class = "factor2", x = match(from@.Data, levels), levels = levels))
          }
    )
    as(object = c("A", "B", "B"), Class = "factor2")
    ```

# Converting from S3 so S4

- It's possible to register an S3 class to be an S4 class by `setOldClass()`. You do this after calling `setClass()`.

    ```{r}
    setClass("factor",
      contains = "integer",
      slots = c(
        levels = "character"
      ),
      prototype = structure(
        integer(),
        levels = character()
      )
    )
    setOldClass("factor", S4Class = "factor")
    ```

- Above, we inherit from the `integer` base type. 

- Notice that we used `structure()` in the prototype, not `list()`. This is because we are building on factors, which are integers with attributes, not list-like objects.

- When you inherit from a base type, it has a slot `.Data` which is where the base type's values live. Let's use this to make new S4 factors.

    ```{r}
    x <- new(Class = "factor", .Data = c(1, 2, 1, 2, 1), levels = c("A", "B"))
    sloop::otype(x)
    x
    x@.Data
    ```

- It is possible to convert any S3 generic to an S4 generic by simply running
    ```{r}
    setGeneric("mean")
    ```

# Method Dispatch

## Multiple inheritance 

- It's possible to create a class that inherits from two or more classes (i.e. have two superclasses). This is called **multiple inheritance**.

- Issue: If the subclass does not not have a method, but both superclasses have methods, then R will choose the method based on alphabetical order. This is dangerous, so we won't cover it.

## Multiple dispatch

- A method can be determined by the classes of multiple objects in the generic call. This is called **multiple dispatch**.

- E.g. how should `c(x, y)` behave if
    1. `x` and `y` are both factors?
    2. `x` is a factor and `y` is a character?
    3.  `x` and `y` are both characters?

- We will demonstrate this using the `Rational` class defined next.


# Rational Vector Example

- I will create a new `Rational` class that contains numerator and denominator slots:

    ```{r}
    setClass(
      Class = "Rational", 
      slots = c(
        num = "integer",
        den = "integer"
      ),
      prototype = list(
        num = integer(),
        den = integer()
      )
    )
    
    setValidity(
      Class = "Rational", 
      method = function(object) {
        if (length(object@num) != length(object@den)) {
          return("numerator length must equal denominator length")
        } else {
          return(TRUE)
        }
      }
    )
    ```

- Let's create a helper function
    ```{r}
    Rational <- function(num = integer(), den = integer()) {
      num <- as.integer(num)
      den <- as.integer(den)
      return(new(Class = "Rational", num = num, den = den))
    }
    ```

- And let's create `show()`, `den()`, and `num()` methods

    ```{r}
    setMethod(
      f = "show", 
      signature = "Rational", 
      definition = function(object) {
        cat(paste0(object@num, "/", object@den))
      }
    )
    
    setGeneric(name = "den", def = function(x) standardGeneric("den"))
    setGeneric(name = "den<-", def = function(x, value) standardGeneric("den<-"))
    setGeneric(name = "num", def = function(x) standardGeneric("num"))
    setGeneric(name = "num<-", def = function(x, value) standardGeneric("num<-"))
    
    setMethod(
      f = "den", 
      signature = "Rational", 
      definition =  function(x) {
        return(x@den)
      }
    )
    
    setMethod(
      f = "num", 
      signature = "Rational", 
      definition =  function(x) {
        return(x@num)
      }
    )
    
    setMethod(
      f = "den<-", 
      signature = "Rational", 
      definition =  function(x, value) {
        x@den <- as.integer(value)
        validObject(x)
        return(x)
      }
    )
    
    setMethod(
      f = "num<-", 
      signature = "Rational", 
      definition =  function(x, value) {
        x@num <- as.integer(value)
        validObject(x)
        return(x)
      }
    )
    ```

- Let's see how it works
    ```{r}
    x <- Rational(c(1, 2, 3), c(9, 5, 2))
    x
    den(x)
    num(x)
    den(x) <- 11:13
    x
    ```
    
# Methods for Group Generic Functions

- Can we add rational objects to each other?
    ```{r, error = TRUE}
    x <- Rational(c(1, 2, 3), c(9, 5, 2))
    y <- Rational(c(5, 2, 6), c(11, 1, 2))
    x + y
    ```

- `+` is not an S3 generic, it is a group generic (see `?S4groupGeneric` for a full list of group generics used by S4).

- S4 automatically allows for group generics to be used as S4 generics, and so we do not need to do that conversion via `setGeneric("+")`. 

- `+` takes arguments `e1` and `e2`, which we will need to use when creating a method:
    ```{r}
    args(getGeneric("+"))
    ```

- Let's create a method for `Rational` addition.

    ```{r}
    setMethod(
      f = "+", 
      signature = "Rational", 
      definition = function(e1, e2) {
        num <- e1@num * e2@den + e2@num * e1@den
        den <- e1@den * e2@den
        return(new(Class = "Rational", num = num, den = den))
      }
    )
    ```

- Now let's try it out
    ```{r}
    x + y
    ```

- Now `1` is just `9/9` or `5/5` etc..., But we cannot just add an integer to a `Rational`?

    ```{r, error = TRUE}
    x + 1L
    ```

- We need to define a method based on **double dispatch**, where one input is a `Rational` and one is an `integer`.
    - In the signature, you tell R which class each argument is.
    - You need to create two methods if you want commutivity. But the second one just calls the first one.

    ```{r}
    setMethod(f = "+", 
      signature = c(e1 = "Rational", e2 = "integer"), 
      definition = function(e1, e2) {
        num <- e1@num + e1@den * e2
        den <- e1@den
        return(new(Class = "Rational", num = num, den = den))
      }
    )
    
    setMethod(f = "+", 
      signature = c(e1 = "integer", e2 = "Rational"), 
      definition = function(e1, e2) e2 + e1
    )
    ```

- It works!
    ```{r}
    x + 2L
    2L  + x
    ```

- A shorthand to define all arithmetic operations is by the `Arith` generic

    ```{r}
    setMethod(
      f = "Arith", 
      signature = c(e1 = "Rational", e2 = "integer"), 
      definition = function(e1, e2) {
        op <- .Generic[[1]]
        if (op == "+") {
          num <- e1@num + e1@den * e2
          den <- e1@den
        } else if (op == "*") {
          num <- e1@num * e2
          den <- e1@den
        } else {
          stop("only +/* are defined right now")
        }
        return(new(Class = "Rational", num = num, den = den))
      }
    )
    
    setMethod(
      f = "Arith",
      signature = c(e1 = "integer", e2 = "Rational"),
      definition = function(e1, e2) {
        op <- .Generic[[1]]
        switch(op,
               "+" = e2 + e1,
               "*" = e2 * e1,
               stop("only +/* are defined right now")
        )
      }
    )
    
    x <- Rational(num = c(1L, 2L, 3L), den = c(4L, 5L, 6L))
    x + 1L
    x * 2L
    2L * x
    ```

- Above, `.Generic` is a variable defined inside the evaluation environment of the `setMethod()` definition. It contains the arithmetic operation applied by the user.

# Documenting S4 objects

- To document an S4 class, add `{roxygen2}` comments before `setClass()`.

- Use the `@slot` tag to document each slot, just like you would document each parameter of a function.

- Make sure you also provide a `@name` tag of the form `classname-class`

- Let's document the `Person` class.
    ```{r, eval = FALSE}
    #' An S4 class representing a person
    #' 
    #' @slot name The name of the person.
    #' @slot age The age of the person.
    #' 
    #' @name Person-class
    setClass(
      Class = "Person",
      slots = c(
        name = "character",
         age = "numeric"
      ),
      prototype = list(
        name = character(),
        age = double()
      )
    )
    ```
    
- S4 generics are regular functions, so document them like usual.

- You can use `@describeIn` to simplify documentation of S4 methods by combining them with the generic or the class documentation.

- In an R package, the files are loaded in alphabetical order. However, for S4 objects, you need to super superclasses before you are allowed to create subclasses.

- To force another file to load before the current file, put this at the top of the current file

    ```{r}
    #' @include foo.R bar.R
    NULL
    ```

- The above says that foo.R and bar.R should be loaded before the current file.

# Some Example Uses of S4 by Others

## `{Matrix}`

- The `{Matrix}` package provides classes for many types of matrices:
    - sparse (mostly zeros)
    - triangular (either the upper or lower triangle elements are all 0).]
    - symmetric (upper and lower triangle elements are equal).
    
    ```{r, message = FALSE}
    library(Matrix)
    ```
    
- Many matrix operations can be more efficiently executed given knowledge of these structures.

- The `USCounties` matrix is a 3111 by 3111 matrix where an element measures a spatial relationship between counties. It is non-zero only if counties are contiguous.

    ```{r}
    data("USCounties", package = "Matrix")
    ```
    
- This is an S4 object of class `dsCMatrix`, used for handling sparse matrices (only 0.0094\% of the values in this matrix are non-zero).

    ```{r}
    sloop::otype(USCounties)
    class(USCounties)
    ```

- Almost every matrix operation you can imagine has been optimized for this sparsity:
    ```{r}
    nrow(sloop::s4_methods_class("dsCMatrix"))
    ```

- This makes things like matrix multiplication much faster:

    ```{r}
    densemat <- as.matrix(USCounties)
    ones <- rep(1, length.out = ncol(densemat))
    microbenchmark::microbenchmark(
      densemat %*% ones,
      USCounties %*% ones
    )
    ```

    - This application doesn't really demonstrate its abilities. It can be even more orders of magnitude faster.

## `{stats4}`

- One application of S4 is in the `{stats4}` package for likelihood inference.

- The **likelihood** is the kind of like the probability of the data that we observed given some parameter values. (not exactly for continuous data)

- E.g., let's say we observed heights (in inches)

    ```{r}
    x <- c(72, 66, 74, 64, 70)
    ```

    and let's assume that these data were generated from a normal distribution.
    
- Then, given any two parameters of the mean (`mu`) and standard deviation (`sigma`), the likelihood of these parameters at each observation would be from `dnorm()`

    ```{r, eval = FALSE}
    dnorm(x = x, mean = mu, sd = sigma)
    ```
    
- If the individuals that these heights are from are **independent** (e.g. they aren't family members), then the likelihood of these parameters given all of the data would be
    ```{r, eval = FALSE}
    prod(dnorm(x = x, mean = mu, sd = sigma))
    ```

- We can calculate this likelihood for a few values of `mu` and `sigma`

    ```{r}
    mu <- 72
    sigma <- 3
    prod(dnorm(x = x, mean = mu, sd = sigma))
    
    mu <- 70
    sigma <- 4
    prod(dnorm(x = x, mean = mu, sd = sigma))
    ```

- Typically, the parameters are not known (`mu` and `sigma` are unknown) and our goal is to estimate them when we only see `x`.

- One way to do this is to find the `mu` and `sigma` that maximize this likelihood. The resulting estimators are called the **maximum likelihood estimators** (MLE's). This is one of the most common methods for estimating parameters from a statistical model.

- Intuition: the MLE's are the parameters that make our observed data as probable to have observed as possible.

- Typically, the likelihood produces very small numbers, so folks maximize the log-likelihood, which results in the same MLE's since $\log()$ is a monotone increasing function.
    ```{r}
    ll <- sum(dnorm(x = x, mean = mu, sd = sigma, log = TRUE)) ## log likelihood
    like <- prod(dnorm(x = x, mean = mu, sd = sigma)) ## likelihood
    
    ## they are the same
    log(like)
    ll
    ```

- The `{stats4}` package provides a bunch of S4 objects and methods for dealing with MLE's.

    ```{r}
    library(stats4)
    ```

- To use the `{stats4}` package, first create a function that takes as input your parameters nothing else. It assumes that all data comes from the parent environment. It should output the *negative* log-likelihood.

    ```{r}
    #' @param mu The mean
    #' @param sigma The standard deviation
    neg_ll <- function(mu, sigma) {
      return(-sum(dnorm(x = x, mean = mu, sd = sigma, log = TRUE)))
    }
    ```

- Then, use the `mle()` function. Include some starting values, and place some limits on constrained parameters.

    ```{r}
    mlout <- stats4::mle(minuslogl = neg_ll, 
                         start = c(mu = 72, sigma = 1), 
                         lower = c(mu = -Inf, sigma = 0),
                         upper = c(mu = Inf, sigma = Inf))
    ```

- This object is S4.
    ```{r}
    sloop::otype(mlout)
    class(mlout)
    ```

- What generics are available for the objects of type `mle`?
    ```{r}
    sloop::s4_methods_class("mle")
    ```

- `coef()` will get you the parameter estimates

    ```{r}
    coef(mlout)
    ```

- `vcov()` will you you the estimated variance/covariance matrix of the MLE's (for their sampling distribution).
    ```{r}
    vcov(mlout)
    ```
    
- So the standard errors are from

    ```{r}
    sqrt(diag(vcov(mlout)))
    ```
    
- `confint()` will get you confidence intervals
    ```{r}
    confint(mlout)
    ```

- In the normal case, these estimates are the same as the sample mean and the sample variance (not adjusting for degrees of freedom).

    ```{r}
    mean(x)
    sqrt(var(x) * (length(x) - 1) / length(x))
    ```

- But the MLE is general to many more complicated applications.

- **Exercise**: The number of heads in $n$ coin flips is binomially distributed with size $n$ and success probability $p$, where $p$ is the probability that a single flip turns up heads. The likelihood, given that `x` heads turned up, is
    ```{r, eval = FALSE}
    dbinom(x = x, size = n, prob = p)
    ```
    
    When $x = 11$ and $n = 31$, calculate the MLE of $p$.

    ```{r, eval = FALSE, echo = FALSE}
    x <- 11
    n <- 31
    neg_ll <- function(pr) {
      -dbinom(x = x, size = n, prob = pr)
    }
    mlout <- mle(minuslogl = neg_ll, 
                 start = c(pr = 0.5), 
                 lower = c(pr = 0), 
                 upper = c(pr = 1))
    coef(mlout)
    ```


# New Functions

- `methods::setClass()`: Define a new S4 class.
- `methods::new()`: Create a new object of a given S4 class.
- `methods::is()`: See the class of an object (if you give it just the object), or test that an object is a certain class (if you give it the object and a class name).
- `methods::slot()`: Extract the data from a slot. Equivalent to `@` notation.
- `methods::setGeneric()`: Create a new S4 generic function
- `methods::setMethod()`: Create a method for a S4 generic.
- `methods::setValidity()`: Enforce structure on an S4 object created by `new()`.
- `methods::validObject()`: Check validity of an S4 object based on `setValidity()` definition.
- `methods::setOldClass()`: Register an S3 class to be an S4 class.
- `methods::setAs()`: Define a coercion method.
- `methods::as()`: Apply a coercion method.
- `methods::getGeneric()`: Get generic function definition.
- `methods::getMethod()`: Get method function definition.
- `base::args()`: Show arguments of a function.
- `sloop::s4_methods_class()`: Show methods of an S4 class.
- `sloop::s4_methods_generic()`: Show methods of an S4 generic.
