---
title: "The `{vctrs}` Package for S3 Vectors"
author: "David Gerard"
date: "`r Sys.Date()`"
output:  
  html_document:
    toc: true
    toc_depth: 4
    toc_float: false
    highlight: pygments
urlcolor: "blue"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo       = TRUE, 
                      fig.align  = "center",
                      fig.height = 3, fig.width = 4)
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(strip.background = ggplot2::element_rect(fill = "white")))
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy()
```

# Learning Objectives

- [`{vctrs}`](https://vctrs.r-lib.org/articles/s3-vector.html) package.
- Building new vector classes.

# Motivation

- What if you want vectors that have special functionality for certain generics. Hadley has a great list:
    - Percent: a double vector that prints as a percentage. 
    - Decimal: a double vector that always prints with a fixed number of decimal places.
    - Cached sum: a double vector that caches the total sum in an attribute.
    - Rational: a pair of integer vectors that defines a rational number like 2 / 3.
    - Polynomial: a list of integer vectors that define polynomials like 1 + x - x^3.
    - Meter: a numeric vector with meter units.
    - Period and frequency: a pair of classes represent a period, or it’s inverse, frequency.
    
- We need to define special S3 classes, along with a variety of useful methods for this class.

- We will learn how to do this using the `{vctrs}` package here.

# Basics

- You create all objects for a given class with `vctrs::new_vctr()`. 

    ```{r}
    x <- vctrs::new_vctr(.data = c(10, 1, 2), class = "percent")
    x
    ```

- The `.data` argument is a vector of one of the base types.

- The `class` argument is the name of the class.

- You should always use `vctrs::new_vctr()` in a constructor function.

    ```{r}
    new_percent <- function(x = double()) {
      stopifnot(is.double(x))
      return(vctrs::new_vctr(x, class = "percent"))
    }
    x <- new_percent(x = c(0.1, 0.01, 0.2))
    x
    ```

- The resulting class inherits from the `vctrs_vctr` class:
    ```{r}
    sloop::s3_class(x)
    ```
    
- When you inherit from the `vctrs_vctr` class, you don't need to deal with the following methods, because they should work automatically:
    - Subsetting: `[`, `[[`, `$` should all return the correct type.
    - `as.data.frame()` will correctly allow a vector to be incorporated in a data frame.
    - Default subset assignment methods will coerce values to match the existing vector (`[<-`, `[[<-`, and `$<-`).

    ```{r}
    x[1]
    x[3] <- 0.55
    x
    data.frame(p = x)
    ```

- Also, you only need to create a `format()` method to print since a `vctrs_vctr` object uses that for `print()` and `str()`.

- To get the underlying `.data` (which you will need to do in the following methods), use `vctrs::vec_data()`.

    ```{r}
    vctrs::vec_data(x)
    ```


- For the helper function, it is nice to try to coerce a user's input to the base type. `vec_cast()` does this more safely than usual.

    ```{r}
    percent <- function(x = double()) {
      x <- vctrs::vec_cast(x, double())
      return(new_percent(x))
    }
    percent(c(1L, 0L))
    ```

- As usual, we should have a type predicate for a new S3 class

    ```{r}
    is_percent <- function(x) {
      inherits(x, "percent")
    }
    is_percent(x)
    ```

# `format()` generic

- The `format()` generic is used for printing nicely.

- `vctrs_vctr` calls `format()` in its `print()` and `str()` generics, so you should always include a `format()` method when using the `{vctrs}` package.

    ```{r}
    format.percent <- function(x, ...) {
      ret <- formatC(vctrs::vec_data(x) * 100, digits = 1, format = "f")
      ret[is.na(x)] <- NA
      ret[!is.na(x)] <- paste0(ret[!is.na(x)], "%")
      return(ret)
    }
    ```

- Now the print method should be altered
    ```{r}
    x <- new_percent(c(0.1, 0.2, NA))
    x
    ```
    
# Casting and Coercion

- **Coercion** is changing the type implicitly. E.g. when we do
    ```{r}
    TRUE + FALSE
    ```

- **Casting** is when we change the type explicitly. E.g. when we do
    ```{r}
    vctrs::vec_cast(c(TRUE, FALSE), double())
    ```

- `{vctrs}` makes casting and coercian more formal.

- You set possible coercions by `vctrs::vec_ptype2()`. This is a generic function.

- You set possible castings using `vctrs::vec_cast()`. This is a generic function.

- Right now, combining vectors to percents is not allowed
    ```{r, error = TRUE}
    vctrs::vec_c(x, 1)
    ```

- I used `vctrs::vec_c()` above because `c()` will never allow us to 

- To allow for coercion, follow these steps:
    1. Create `vec_ptype2.class1.class2()` and `vec_ptype2.class2.class1()` methods. These should only return an empty vector of the class that should result.
        ```{r}
        vec_ptype2.percent.double <- function(x, y, ...) {
          percent()
        }
        
        vec_ptype2.double.percent <- function(x, y, ...) {
          percent()
        }
        ```
    2. Verify that you implemented the coercion correctly with `vec_ptype_show()`
        ```{r}
        vctrs::vec_ptype_show(percent(), double(), percent())
        ```
        - The above says that percent with double and double with percent will return a percent
    3. Create `vec_cast.class1.class2()` and `vec_cast.class2.class1()` methods for casting from the first class to the second class
        ```{r}
        # from double to percent
        vec_cast.percent.double <- function(x, to, ...) percent(x) 
        # from percent to double
        vec_cast.double.percent <- function(x, to, ...) vctrs::vec_data(x) 
        ```
    4. Verify that your casting works with with `vec_cast()`.
        ```{r}
        vctrs::vec_cast(percent(0.5), double()) # from percent to double
        vctrs::vec_cast(0.5, percent()) # from double to percent
        ```
    5. Verify that methods now work
        ```{r}
        vctrs::vec_c(1, percent(0.5))
        ```
        - Note that `c()` does not work
            ```{r}
            c(1, percent(0.5))
            ```
    
- Note that it is usually better to explicitly state that the percent can combine with the percent, because it is faster and allows for different attributes between values.

```{r}
x <- percent(c(0.1, 0.2, 0.1))
y <- percent(0.5)

## allowed
c(x, y)

## Benchmark
microbenchmark::microbenchmark(c(x, y))

# Implement coercion and casting within same class
vec_ptype2.percent.percent <- function(x, y, ...) percent()
vec_cast.percent.percent <- function(x, to, ...) x

# Faster
microbenchmark::microbenchmark(c(x, y))
```

# New Functions

- `vctrs::new_vctr()`: Create a vector object of a given class which inherits from the `vctrs_vctr` class.
- `vctrs::vec_cast()`: Coerce an object to a given type.
- `vctrs::vec_data()`: Get underlying data from a `vctrs_vctr` object.
- `vctrs::vec_ptype2()`: Generic function for setting up possible coercions.
- `vctrs::vec_c()`: Combine vectors with proper coercion.


