---
title: "Performant R Code"
author: "David Gerard"
date: "`r Sys.Date()`"
output:  
  html_document:
    toc: true
    toc_depth: 4
    toc_float: false
    highlight: pygments
urlcolor: "blue"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo       = TRUE, 
                      fig.align  = "center",
                      fig.height = 3, fig.width = 4)
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(strip.background = ggplot2::element_rect(fill = "white")))
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy()
```

# Learning Objectives

- Overview of Metaprogramming.
- Learn a couple tricks.
- Chapter 24 of [Advanced R](https://adv-r.hadley.nz/).
- Strategies for faster R code.

# Motivation

- Before you get started, the **two rules of software optimization** are [@jackson1975principles]:
    1. Don't do it.
    2. Don’t do it yet (for experts only).

# Keep Multiple Approaches

- Your first goal should be clarity and robustness, not speed.

- Write code in an unoptimized but clear.

- Once you have it working and you want to optimize, do not delete that approach. 

- Rather, write a new function with the new approach, and compare the old and new approaches in a unit test.

# Use Less General Functions

- Using functions that are specific to a problem is almost always faster than using functions that are general.

- Using `rowSums()` (very specific) is faster than `apply()` (very general).
- Using `vapply()` (prespecifies output so more specific) is faster than `sapply()` (does not prespecify output).
- `any(x == 10)` is faster than `10 %in% x` because testing for equality is more specific than testing for inclusion.

# Vectorize

- Vectorization means operating on whole vectors instead of individual elements of a vectors.

- E.g., doing

    ```{r}
    x <- 1:100
    x <- x + 1
    ```
   
   instead of 
   
    ```{r}
    x <- 1:100
    for (i in seq_along(x)) {
      x[[i]] <- x[[i]] + 1
    }
    ```
   
- It is almost always faster to vectorize because a lot of those underlying implementations are written in C (which is faster).

- Whenever you use a for-loop, first think if there is another way that you can do it without the for-loop. 

- I don't mean try using `map()` or `lapply()`, since those aren't much faster than for-loops.

- I mean try to use base operations where possible.

- Use subassignment:
    ```{r, eval = FALSE}
    x[is.na(x)] <- 0
    ```

- Use look-up tables
    ```{r}
    x <- c(a = "alive", d = "dead")
    y <- sample(c("a", "d"), 10, TRUE)
    x[y]
    ```

- Use R's linear algebra operations. Whenever you can convert operations to linear algebra, it tends to run super fast since there are lots of optimizations that occur in the C libraries R uses.

# Avoid Copies

- Pre-define your outputs before you fill them up in a for-loop.

- Copy-on-modify is very expensive. You can only avoid it if you pre-define your outputs.

- If you cannot pre-define your outputs, use a list, since lists just make shallow copies when they copy-on-modify.

# Write in C++

- If you need to use a for-loop, and you cannot vectorize, that is when you shift to C.

# References
