---
title: "Batch Processing and LSF"
author: "David Gerard"
date: "`r Sys.Date()`"
output:  
  html_document:
    toc: true
    toc_depth: 4
    toc_float: false
    highlight: pygments
urlcolor: "blue"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo       = TRUE, 
                      fig.align  = "center",
                      fig.height = 3, fig.width = 4)
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(strip.background = ggplot2::element_rect(fill = "white")))
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy()
```

# Learning Objectives

- Batch Processing
- Schedulars
- LSF ("load sharing facility")

# Running R in Batch

- `R CMD BATCH`

# LSF

- American University has a supercomputer called [Zorro](https://www.american.edu/cas/hpc/)

- To use it, you have to submit jobs via a [schedular](https://en.wikipedia.org/wiki/Job_scheduler).

- The schedular Zorro uses is [LSF](https://en.wikipedia.org/wiki/IBM_Spectrum_LSF), so we'll learn that.

- There are many other, popular schedulars. The syntax is all different, but they all operate on similar ideas.

# New Functions

