---
title: "Batch Processing and LSF"
author: "David Gerard"
date: "`r Sys.Date()`"
output:  
  html_document:
    toc: true
    toc_depth: 4
    toc_float: false
    highlight: pygments
urlcolor: "blue"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo       = TRUE, 
                      fig.align  = "center",
                      fig.height = 3, fig.width = 4)
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(strip.background = ggplot2::element_rect(fill = "white")))
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy()
```

# Learning Objectives

- Batch Processing
- Schedulars and LSF ("load sharing facility")
- [LSF bsub options](https://www.ibm.com/docs/en/spectrum-lsf/10.1.0?topic=bsub-options)
- [Minimal LSF Example](https://github.com/dcgerard/minmin)

# Running R in Batch

- You can run an R script file in [batch mode](https://en.wikipedia.org/wiki/Batch_processing) (i.e. in the background) from bash by using the following command:

    ```{bash, eval = FALSE}
    R CMD BATCH --no-save --no-restore input_file.R output_file.Rout
    ```
    
    Make sure to change "input_file.R" and "output_file.Rout"

    - The `--no-save --no-restore` options make sure that you are working with a clean environment and that you don’t save this environment after the command is executed. This is a good thing for reproducibility.

- The newer way is to use `Rscript`. The following command will do mostly what `R CMD BATCH` will do, except only the output of `print()` will be sent to "output_file.Rout", and the console output will not be sent.

    ```{bash, eval = FALSE}
    Rscript input_file.R > output_file.Rout
    ```
    
- Even though `R CMD BATCH` is an older way to run R scripts in batch mode, I like it better because:
    1. `Rscript` forces the `--no-echo` option, so console output is not printed.
    2. I've had stability issues with `Rscript` when using it on the super computer. But that was a few years ago and they might have fixed things by now.
    
- One nice thing about `Rscript` is that you can run one-liners of R code from bash using `-e`:

    ```{bash}
    Rscript -e 'print("hello world")'
    ```

- This allows you to knit an R markdown file from bash by using the following command:

    ```{bash, eval = FALSE}
    Rscript -e "library(rmarkdown);render('rmarkdown_file.Rmd')"
    ```

- On Windows machines, the above commands should also work on [Windows PowerShell](https://en.wikipedia.org/wiki/PowerShell), but not Git Bash for Windows.

- **Exercise**: Write an R script that:
    1. Loads `mtcars`
    2. Makes a scatterplot of `mpg` on `wt`
    3. Saves the output to a file called "plot.pdf"
    
    Then run this script in batch mode.

# Executable R Scripts

- On Unix-like machines (e.g. Ubuntu and Mac), you can make an R file executable using [shebang](https://en.wikipedia.org/wiki/Shebang_(Unix)).

- At the very top of your R script, place the following
    ```{r, eval = FALSE}
    #!/path/to/Rscript --vanilla
    ```
    where "/path/to/Rscript" is replaced with where Rscript is installed on your machine.
    
    - The `--vanilla` option makes sure that your script is reproducible in other users' computing environments.
    
- You can find where `Rscript` is located by running the following in bash:

    ```{bash}
    which Rscript
    ```

- It is common for "/usr/bin/Rscript" to be your `Rscript` location, unless you changed the defaults.
    
- Then change the mode to executable using [`chmod`](https://en.wikipedia.org/wiki/Chmod)

    ```{bash, eval = FALSE}
    chmod +x filename.R
    ```

- Then you can execute it by running the following in bash
    ```{bash, eval = FALSE}
    ./filename.R
    ```

- You can make it so that you can execute this file anywhere on your computer by [adding it to the `$PATH`](https://stackoverflow.com/questions/14637979/how-to-permanently-set-path-on-linux-unix).

- This makes creating [command line interfaces](https://en.wikipedia.org/wiki/Command-line_interface) for your R code relatively simple.

- Suppose I have the following R script file called "ex1.R"

    ```{r, eval = FALSE}
    #!/usr/bin/Rscript --vanilla
    cat("Hello World\n", file = "hello.txt")
    ```

- Then I run
    ```{bash, eval = FALSE}
    chmod +x ex1.R
    ```

- Then, whenever I run the following, that script will run:
    ```{bash, eval = FALSE}
    ./ex1.R
    ```

# LSF

- American University has a supercomputer called [Zorro](https://www.american.edu/cas/hpc/).

- To use it, you have to submit jobs via a [schedular](https://en.wikipedia.org/wiki/Job_scheduler).

- The schedular Zorro uses is [LSF](https://en.wikipedia.org/wiki/IBM_Spectrum_LSF), so we'll learn that.

- There are many other, popular schedulars. The syntax is all different, but they all operate on similar ideas.

## Preparations

- You need faculty sponsorship to get access to Zorro. Once you have a sponsorship, fill out a request form from the [Zorro Website](https://www.american.edu/cas/hpc/).

- After you have permission to access Zorro, you interface with it using [SSH](https://en.wikipedia.org/wiki/Secure_Shell).
    - Windows users should download and install [PuTTY](https://www.chiark.greenend.org.uk/~sgtatham/putty/latest.html) to use SSH.
    - Mac users already have SSH installed in bash.

- You transfer files between the supercomputer and your computer through [FTP](https://en.wikipedia.org/wiki/File_Transfer_Protocol).
    - Windows users should download and install [WinSCP](www.winscp.net) to do use FTP. I sometimes also use [FileZilla](https://filezilla-project.org/)
    - Mac users can already use FTP through the terminal. But using [FileZilla](https://filezilla-project.org/) might be a little easier.

- You are required to use the AU VPN while accessing Zorro. All users should follow the AU instructions [here](https://www.american.edu/oit/network/vpn.cfm) to connect to the AU VPN.

## Connecting to Zorro

- Connect to the [AU VPN](https://www.american.edu/oit/network/vpn.cfm).

- Windows Users: 
    1. Open PuTTY
    2. Under Host Name, type `zorro.american.edu` to connect to the server. Press Enter
    3. A black screen will appear where you need to enter your AU username and password.
    
- Mac Users:
    1. In the terminal, type:
        ```{bash, eval = FALSE}
        ssh -l username zorro.american.edu
        ```
        where "username" is your AU username.
    2. Type in your AU password and hit enter.
    
- Once you are connected to Zorro, you use [bash](./01_basic_bash.html) to navigate, run programs, submit jobs, etc..

## LSF Files

- You format the properties of a job in terms of a LSF file.

- Example LSF File:

    ```{bash, eval = FALSE}
    #BSUB -J minimal_example
    #BSUB -q normal
    #BSUB -o minimal_out.txt
    #BSUB -e minimal_err.txt
    #BSUB -u "youremail@american.edu"
    #BSUB -B
    #BSUB -N
    #BSUB n=2
    /path/to/R CMD BATCH --no-save --no-restore '--args nc=2' minimal.R minimal.Rout
    ```

- You have a list of LSF options. Each option begins with `#BSUB`

- You then have a bash command. In this case we are running R in batch mode via
    ```{bash, eval = FALSE}
    /path/to/R CMD BATCH --no-save --no-restore '--args nc=2' minimal.R minimal.Rout
    ```
    
- The "path/to/R" can be found by typing the following in bash:
    ```{bash, eval = FALSE}
    which R
    ```

- This might not be the version of R that you want. Zorro has a few copies or R installed. You can see these versions by typing in bash:

    ```{bash, eval = FALSE}
    ls -d /app/R-*
    ```

- As of the writing of this document, they have versions 3.6.0, 3.6.1, 4.0.2, and 4.1.0
    
- You can use a specific version of R (e.g. 4.0.2) by using

    ```{bash, eval = FALSE}
    /app/R-4.0.2/bin/R CMD BATCH --no-save --no-restore '--args nc=2' minimal.R minimal.Rout
    ```

- That `'--args nc=2'` trick is a way to pass arbitrary arguments to an R script. See [here](https://www.r-bloggers.com/2007/08/including-arguments-in-r-cmd-batch-mode/).

- Typical options are:
    - `#BSUB -J job_name`: Name of the job, so that you can see that name when you check the status of your job.
    - `#BSUB -q normal`: The queue to submit your job to. You shouldn't change this for Zorro.
    - `#BSUB -o job_out.txt`: Where to send the job's output.
    - `#BSUB -e job_err.txt`: Where to send the job's error messages.
    - `#BSUB -u "youremail@american.edu"`: Send mail to the specified user.
    - `#BSUB -B`: Emails you when the job begins.
    - `#BSUB -N`: Emails you when the job ends.
    - `#BSUB n=2`: Submits a parallel job and specifies the number of tasks in the job (in this case, 2).

- A full list of options can be found [here](https://www.ibm.com/docs/en/spectrum-lsf/10.1.0?topic=bsub-options)


## Paired R script

- You need to format your R script to be able to run in parallel.

- Example R Script

    ```{r, eval = FALSE}
    ## Set library for R packages ----
    .libPaths(c("/home/dgerard/R/4.0.2/", .libPaths()))
    
    ## Attach packages for parallel computing ----
    library(foreach)
    library(doFuture)
    
    ## Determine number of cores ----
    args <- commandArgs(trailingOnly = TRUE)
    if (length(args) == 0) {
      nc <- 1
    } else {
      eval(parse(text = args[[1]]))
    }
    cat(nc, "\n")
    
    ## Register workers ----
    if (nc == 1) {
      registerDoSEQ()
      plan(sequential)
    } else {
      registerDoFuture()
      plan(multisession, workers = nc)
      if (getDoParWorkers() == 1) {
        stop("nc > 1, but only one core registered")
      }
    }
    
    ## Run R script ----
    x <- foreach(i = 1:2, .combine = c) %dopar% {
      Sys.sleep(1)
      i
    }
    x
    
    ## Unregister workers ----
    if (nc > 1) {
      plan(sequential)
    }
    ```

- The above is the template I use.

- The above only allows to run on multiple cores on one node. So the above does not allow for multi-node processing.

- You should only need to modify two things in the above code:
    1. The code right after `## Run R script ----`, where you implement the your computations.
    2. The path for your R library, in the `.libPaths()` call. This is where your R packages are locally stored (see below).

## `bsub` commands

- You submit and control jobs with the `bsub` command in bash.

- The following will submit the job in `"minimal.lsf"`

    ```{bash, eval = FALSE}
    bsub < minimal.lsf
    ```

    - The above says to pipe the `minimal.lsf` file to the bsub command. 
    - Otherwise, you would need to use `bsub` like `bsub -q normal -n 2 ...` 

- Use `bjobs` to display information on the jobs that you have submitted.
    - Display all of your jobs:
        ```{bash, eval = FALSE}
        bjobs -a
        ``` 
    - Display all of the jobs of a user:
        ```{bash, eval = FALSE}
        bjobs -u user_name
        ```
    - Display information about a particular job
        ```{bash, eval = FALSE}
        bjobs job_id
        ```

- Different job states that you will see:
    - `PEND`: Waiting in a queue.
    - `RUN`: Currently running.
    - `DONE`: Successfully finished with no errors.
    - `EXIT`: Errored, did not finish successfully.

- Use `bkill` to kill a job.
    ```{bash, eval = FALSE}
    bkill job_id
    ```

## Installing R Packages

- You can see what packages are already installed on Zorro by typing

    ```{bash, eval = FALSE}
    ls /app/R-4.0.2/lib64/R/library
    ```

    - It's not a lot.

- You need to install R packages in a local directory, since global install is not supported (because not everyone wants your R packages).

- You should create an R directory where you put all things R:

    ```{bash, eval = FALSE}
    mkdir R
    ```

- Then inside this directory, create a directory where you can place your packages for a specific version of R

    ```{bash, eval = FALSE}
    mkdir 4.0.2
    ```

- Then create the following R file, called "install.R", which can be used to install R packages into that directory (but change my username to yours):
    ```{r, eval = FALSE}
    .libPaths(c("/home/dgerard/R/4.0.2/", .libPaths()))
    install.packages(c("tidyverse", 
                       "future",
                       "doFuture",
                       "foreach"),
                     lib = "/home/dgerard/R/4.0.2/",
                     repos = "http://cran.us.r-project.org")
    ```
    - In the above, I install `{tidyverse}`, `{future}`, `{doFuture}`, and `{foreach}` packages. But you can add more.
    
- If you want to use Bioconductor packages in R Version 4.0.2, run the following:

    ```{r, eval = FALSE}
    .libPaths(c("/home/dgerard/R/4.0.2/", .libPaths()))
    install.packages("BiocManager",
                     lib = "/home/dgerard/R/4.0.2/",
                     repos = "http://cran.us.r-project.org")
    BiocManager::install(version = "3.12", lib = "/home/dgerard/R/4.0.2/", ask = FALSE)
    BiocManager::install(c("tidyverse",
                           "future",
                           "doFuture",
                           "foreach"),
                         lib = "/home/dgerard/R/4.0.2/",
                         ask = FALSE)
    ```
    
    - I got the version of Bioconductor corresponding to R 4.0.2 from [here](https://www.bioconductor.org/about/release-announcements/#release-versions)

- Set up an LSF file to run this script, called "install.lsf":

    ```{bash, eval = FALSE}
    #BSUB -J install_r_pkgs
    #BSUB -q normal
    #BSUB -o install_out.txt
    #BSUB -e install_err.txt
    #BSUB -u "youremail@american.edu"
    #BSUB -B
    #BSUB -N
    #BSUB n=1
    /app/R-4.0.2/bin/R CMD BATCH --no-save --no-restore install.R install.Rout
    ```
    
- Then run this job

    ```{bash, eval = FALSE}
    bsub < install.lsf
    ```

    ```{bash, eval = FALSE, echo = FALSE}
    bsub -q normal R CMD BATCH install.R install_out.txt
    ```

- At the top of every R file from now on, put the following (but change my username to yours):
    ```{r, eval = FALSE}
    .libPaths(c("/home/dgerard/R/4.0.2/", .libPaths()))
    ```

- Now you have access to those packages that you installed in the "R/4.0.2" directory.

# Installing other software locally

- For some R packages, you need to have additional software installed. But global install is not allowed on the super computer.

- Here are the general steps:

1. Download a tar file using `wget <software_url>`

2. Decompress the tar file using `tar -zxvf <tar_file>`

3. Move into the newly decompressed file using `cd <new_file>`

4. Look at the README for further steps. Usually there is a makefile and it's as easy as running `make` and/or `make install`. But read the README first.

5. Add the location to the `PATH` by using something like `export PATH=$HOME/path/to/software:$PATH`

6. Source the `.bash_profile` with `. ~/.bash_profile`

7. Confirm that the software is installed with `which <software>`

- You can see an example of this pipeline [here](https://help.dreamhost.com/hc/en-us/articles/360001435926-Installing-OpenSSL-locally-under-your-username)

- This pipeline won't work all of the time. You need to read the docs.

```{block, eval = FALSE, echo = FALSE}
- This worked to install `sl` locally

yumdownloader --source sl

This downloaded the file "sl-5.02-1.e17.src.rpm"

rpm -i --prefix=/myapps/ sl-5.02-1.e17.src.rpm
```


# New Functions

- `R CMD BATCH`: Run R in batch mode (classic way).
- `Rscript`: Run R in batch mode (newer way).
- `.libPaths()`: Set and get the paths where R will search for installed packages.
- `bsub`: Submit an LSF job.
- `bjobs`: See status of LSF jobs.
- `bkill`: Kill an LSF job.
