---
title: "Batch Processing and LSF"
author: "David Gerard"
date: "`r Sys.Date()`"
output:  
  html_document:
    toc: true
    toc_depth: 4
    toc_float: false
    highlight: pygments
urlcolor: "blue"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo       = TRUE, 
                      fig.align  = "center",
                      fig.height = 3, fig.width = 4)
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(strip.background = ggplot2::element_rect(fill = "white")))
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy()
```

# Learning Objectives

- Batch Processing
- Schedulars and LSF ("load sharing facility")
- [`{future.batchtools}` Vignette](https://cran.r-project.org/web/packages/future.batchtools/vignettes/future.batchtools.html)

# Running R in Batch

- You can run an R script file in [batch mode](https://en.wikipedia.org/wiki/Batch_processing) (i.e. in the background) from bash by using the following command:

    ```{bash, eval = FALSE}
    R CMD BATCH --no-save --no-restore input_file.R output_file.Rout
    ```
    
    Make sure to change "input_file.R" and "output_file.Rout"

    - The `--no-save --no-restore` options make sure that you are working with a clean environment and that you don’t save this environment after the command is executed. This is a good thing for reproducibility.

- The newer way is to use `Rscript`. The following command will do mostly what `R CMD BATCH` will do, except only the output of `print()` will be sent to "output_file.Rout", and the console output will not be sent.

    ```{bash, eval = FALSE}
    Rscript input_file.R > output_file.Rout
    ```
    
- Even though `R CMD BATCH` is an older way to run R scripts in batch mode, I like it better because:
    1. `Rscript` forces the `--no-echo` option, so console output is not printed.
    2. I've had stability issues with `Rscript` when using it on the super computer. But that was a few years ago and they might have fixed things by now.
    
- One nice thing about `Rscript` is that you can run one-liners of R code from bash using `-e`:

    ```{bash}
    Rscript -e 'print("hello world")'
    ```

- This allows you to knit an R markdown file from bash by using the following command:

    ```{bash, eval = FALSE}
    Rscript -e "library(rmarkdown);render('rmarkdown_file.Rmd')"
    ```

- On Windows machines, the above commands should also work on [Windows PowerShell](https://en.wikipedia.org/wiki/PowerShell), but not Git Bash for Windows.

- **Exercise**: Write an R script that:
    1. Loads `mtcars`
    2. Makes a scatterplot of `mpg` on `wt`
    3. Saves the output to a file called "plot.pdf"
    
    Then run this script in batch mode.

# Executable R Scripts

- On Unix-like machines (e.g. Ubuntu and Mac), you can make an R file executable using [shebang](https://en.wikipedia.org/wiki/Shebang_(Unix)).

- At the very top of your R script, place the following
    ```{r, eval = FALSE}
    #!/path/to/Rscript --vanilla
    ```
    where "/path/to/Rscript" is replaced with where Rscript is installed on your machine.
    
    - The `--vanilla` option makes sure that your script is reproducible in other users' computing environments.
    
- You can find where `Rscript` is located by running the following in bash:

    ```{bash}
    which Rscript
    ```

- It is common for "/usr/bin/Rscript" to be your `Rscript` location, unless you changed the defaults.
    
- Then change the mode to executable using [`chmod`](https://en.wikipedia.org/wiki/Chmod)

    ```{bash, eval = FALSE}
    chmod +x filename.R
    ```

- Then you can execute it by running the following in bash
    ```{bash, eval = FALSE}
    ./filename.R
    ```

- You can make it so that you can execute this file anywhere on your computer by [adding it to the `$PATH`](https://stackoverflow.com/questions/14637979/how-to-permanently-set-path-on-linux-unix).

- This makes creating [command line interfaces](https://en.wikipedia.org/wiki/Command-line_interface) for your R code relatively simple.

- Suppose I have the following R script file called "ex1.R"

    ```{r, eval = FALSE}
    #!/usr/bin/Rscript --vanilla
    cat("Hello World\n", file = "hello.txt")
    ```

- Then I run
    ```{bash, eval = FALSE}
    chmod +x ex1.R
    ```

- Then, whenever I run the following, that script will run:
    ```{bash, eval = FALSE}
    ./ex1.R
    ```

# LSF

- American University has a supercomputer called [Zorro](https://www.american.edu/cas/hpc/).

- To use it, you have to submit jobs via a [schedular](https://en.wikipedia.org/wiki/Job_scheduler).

- The schedular Zorro uses is [LSF](https://en.wikipedia.org/wiki/IBM_Spectrum_LSF), so we'll learn that.

- There are many other, popular schedulars. The syntax is all different, but they all operate on similar ideas.

## Preparations

- You need faculty sponsorship to get access to Zorro. Once you have a sponsorship, fill out a request form from the [Zorro Website](https://www.american.edu/cas/hpc/).

- After you have permission to access Zorro, you interface with it using [SSH](https://en.wikipedia.org/wiki/Secure_Shell) and 
    - Windows users should download and install [PuTTY](https://www.chiark.greenend.org.uk/~sgtatham/putty/latest.html) to use SSH.
    - Mac users already have SSH installed in bash.

- You transfer files between the supercomputer and your computer through [FTP](https://en.wikipedia.org/wiki/File_Transfer_Protocol).
    - Windows users should download and install [WinSCP](www.winscp.net) to do use FTP. I sometimes also use [FileZilla](https://filezilla-project.org/)
    - Mac users can already use FTP through the terminal. But using [FileZilla](https://filezilla-project.org/) might be a little easier.

- You are required to use the AU VPN while accessing Zorro. All users should follow the AU instructions [here](https://www.american.edu/oit/network/vpn.cfm) to connect to the AU VPN.

## Connecting to Zorro

- Connect to the [AU VPN](https://www.american.edu/oit/network/vpn.cfm).

- Windows Users: 
    1. Open PuTTY
    2. Under Host Name, type `zorro.american.edu` to connect to the server. Press Enter
    3. A black screen will appear where you need to enter your AU username and password.
    
- Mac Users:
    1. In the terminal, type:
        ```{bash, eval = FALSE}
        ssh -l username zorro.american.edu
        ```
        where "username" is your AU username.
    2. Type in your AU password and hit enter.
    
- Once you are connected to Zorro, you use [bash](./01_basic_bash.html) to navigate, run programs, submit jobs, etc...

# New Functions

- `R CMD BATCH`: Run R in batch mode (classic way).
- `Rscript`: Run R in batch mode (newer way).
