---
title: "Final Project Description"
author: "David Gerard"
date: "`r Sys.Date()`"
output:  
  html_document:
    toc: true
    toc_depth: 4
    toc_float: false
    highlight: pygments
urlcolor: "blue"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo       = TRUE, 
                      fig.align  = "center",
                      fig.height = 3, fig.width = 4)
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(strip.background = ggplot2::element_rect(fill = "white")))
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy()
```

# Summary

You will work in teams of two or three to build an R package using some of the tools we have learned during this course.

# Grading Criteria

- Generic functions used.
- Iteration used.
- C++ functions used.
- Unit testing implemented.
    - Coverage > 70\%
- Continuous integration
    - No notes/warnings/errors
- Assertions
    - Your functions throw errors at impoper inputs.

# Grade breakdown

- Project selection: 1 pt
- Progress report: 5 pts
- Presentation: 5 pts
- R Package: 10 pts

# Project Topics

- You are responsible for choosing a project of sufficient complexity. But here are some possible topics that you may consider.

- I'll update this page as I gain more ideas.

## Small Groups

- A **binary operator** is an operation that combines two elements to create a third element. E.g. $+$ is a binary operator for numbers, we can "combine" 1 and 2 to make 3 via  $1 + 2 = 3$

- A **group** is a space (denoted $G$) of elements along with a binary operator (denoted "$\cdot$") such that
    - **Associativity**: for all $a,b,c\in G$ we have
        $$
        (a \cdot b) \cdot c = a \cdot (b \cdot c)
        $$
    - **Identity**: There exists an $e\in G$ such that
        $$
        e \cdot a = a \cdot e = a
        $$
        for all $a \in G$.
    - **Inverse**: For all $a \in G$ there exists a $b \in G$ such that
        $$
        a \cdot b = b \cdot a = e
        $$
        We usually denote $b = a^{-1}$.

- Groups are fundamental to the building blocks of much of theoretical mathematics.

- Wikipedia provides a list of small groups: <https://en.wikipedia.org/wiki/List_of_small_groups>

- **Project**: Create an S4 object oriented system for some of the small groups from the Wikipedia page on small groups.

- E.g. Suppose that `as.g_4_2()` converts a numeric vector into a vector encoding the [Klein 4-group](https://en.wikipedia.org/wiki/Klein_four-group), and we define `+` to be the binary operator (which you would need to overload). Then I am envisioning code of the form

    ``` r
    x <- as.g_4_2(c(1, 2, 3, 4))
    x + 1
    ```
    
    ```
    # 1
    # 2
    # 3
    # 4
    ```
    
    ``` r
    x + 2
    ```
    
    ```
    # 2
    # 1
    # 4
    # 3
    ```
    
    ``` r
    x + 3
    ```
    
    ```
    # 3
    # 4
    # 1
    # 2
    ```
    
    ``` r
    x + 4
    ```
    
    ```
    # 4
    # 3
    # 2
    # 1
    ```

- A good S4 object system that implements binary operations that you can explore is [`{lubridate}`](https://github.com/tidyverse/lubridate). 
    - E.g. they add `duration`s together in a special way.

## Marvel API

- **Description**: Use the `{httr}` package to build an R package interface to the Marvel Comic's API: <https://developer.marvel.com/>

- I would only recommend this project if you are familiar with the `{httr}` package: <https://data-science-master.github.io/lectures/08_web_scraping/08_apis.html>

- You should create R functions that translate into HTTP queries through `{httr}`. 

- You should follow best practices for API packages as detailed by the `{httr}` vignette: <https://httr.r-lib.org/articles/api-packages.html>

- Your functions should return to the user results in the form of a tidy data frame.

- I would recommend looking at examples from the [`{tidycensus}`](https://cran.r-project.org/package=tidycensus) package

- The API has endpoints for comics, comic series, comic stories, comic events and crossovers, creators, and characters. I would recommend creating a function for each of these endpoints.

- For this project, you can try other API's (<https://github.com/public-apis/public-apis>). I just thought this one looked well-documented and fun.
